---
title: "Comparative co-expression analysis with rcomplex"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparative co-expression analysis with rcomplex}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_se <- requireNamespace("SummarizedExperiment", quietly = TRUE)
has_torch <- requireNamespace("torch", quietly = TRUE) &&
             torch::torch_is_installed() &&
             tryCatch({ torch::torch_randn(1); TRUE }, error = function(e) FALSE)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      eval = has_se && has_torch)

# Helper for interactive tables (sortable by one or multiple columns)
show_table <- function(df, pageLength = 10, ...) {
  if (requireNamespace("DT", quietly = TRUE)) {
    DT::datatable(df, rownames = FALSE,
                  options = list(pageLength = pageLength, scrollX = TRUE,
                                 ordering = TRUE, orderMulti = TRUE),
                  ...)
  } else {
    knitr::kable(df)
  }
}
```

This vignette demonstrates a complete rcomplex analysis using expression
data from eight Pooideae grass species spanning four phylogenetic pairs,
each containing one annual and one perennial species:

| Species | Code | Life cycle | Pair |
|---------|------|------------|------|
| *Brachypodium distachyon* | BDIS | Annual | Brachypodium |
| *Brachypodium sylvaticum* | BSYL | Perennial | Brachypodium |
| *Hordeum vulgare* | HVUL | Annual | Hordeum |
| *Hordeum jubatum* | HJUB | Perennial | Hordeum |
| *Briza maxima* | BMAX | Annual | Briza |
| *Briza media* (crossmapped to BMAX) | BMED | Perennial | Briza |
| *Vulpia bromoides* | VBRO | Annual | Festuca/Vulpia |
| *Festuca pratensis* | FPRA | Perennial | Festuca/Vulpia |

The dataset contains the 1500 most highly expressed HOGs from VST-normalized RNA-seq from the
"Live or let die?" project, stored as SummarizedExperiment objects.

## 1. Data preparation

```{r data, message = FALSE}
library(rcomplex)
library(SummarizedExperiment)

se_list <- readRDS(system.file("extdata", "pooideae_vignette.rds",
                                package = "rcomplex"))
species <- names(se_list)

# Life cycle trait
trait <- c(BDIS = "annual", BSYL = "perennial",
           HVUL = "annual", HJUB = "perennial",
           BMAX = "annual", BMED = "perennial",
           VBRO = "annual", FPRA = "perennial")
```

### Paralog reduction

Multi-copy gene families often contain recent duplicates with nearly
identical expression profiles. These inflate the combinatorial search
space for clique detection without adding information.
`reduce_orthogroups()` merges correlated paralogs within each HOG
using Ward.D2 hierarchical clustering on Pearson correlation distance.
Paralogs above the correlation threshold (default 0.7) are replaced
by their averaged expression; subfunctionalized copies with distinct
expression programs are preserved:

```{r reduce}
# Build a per-species ortholog table from rowData HOG annotations
reduced_se <- lapply(species, function(sp) {
  expr <- SummarizedExperiment::assay(se_list[[sp]], 1)
  rd <- SummarizedExperiment::rowData(se_list[[sp]])
  ortho_sp <- data.frame(
    gene = rownames(se_list[[sp]]),
    hog = as.character(rd$hog),
    stringsAsFactors = FALSE
  )
  ortho_sp <- ortho_sp[!is.na(ortho_sp$hog), ]
  reduced <- reduce_orthogroups(expr, ortho_sp, gene_col = "gene",
                                 cor_threshold = 0.7)
  cat(sp, ":", reduced$n_original, "->", reduced$n_reduced, "genes",
      "(", reduced$n_merged, "merged )\n")
  reduced
})
names(reduced_se) <- species
```

### Network construction

Build co-expression networks from the reduced expression matrices
using Mutual Rank (MR) normalization at 3% density. We pass the
matrix directly since SE metadata is no longer needed post-reduction:

```{r networks}
networks <- lapply(species, function(sp) {
  compute_network(reduced_se[[sp]]$expr_matrix,
                  norm_method = "MR", density = 0.03)
})
names(networks) <- species
```

### Ortholog extraction

Extract ortholog pairs from HOG annotations in rowData. After
paralog reduction, gene names may have changed (merged representatives),
so we map through each species' reduction. Gene IDs are species-unique
(genome-specific identifiers), so each species' map is applied only
to the column containing that species' genes:

```{r orthologs}
ortho_pairs <- utils::combn(species, 2, simplify = FALSE)
ortho_list <- lapply(ortho_pairs, function(pair) {
  o <- extract_orthologs(se_list[[pair[1]]], se_list[[pair[2]]])
  o$sp1_name <- pair[1]
  o$sp2_name <- pair[2]
  o
})
orthologs <- do.call(rbind, ortho_list)

# Map original gene names to reduced representatives (species-aware)
for (sp in species) {
  gm <- reduced_se[[sp]]$gene_map
  map <- stats::setNames(gm$representative, gm$original)
  is_sp1 <- orthologs$sp1_name == sp
  is_sp2 <- orthologs$sp2_name == sp
  orthologs$Species1[is_sp1] <- ifelse(
    orthologs$Species1[is_sp1] %in% names(map),
    map[orthologs$Species1[is_sp1]], orthologs$Species1[is_sp1])
  orthologs$Species2[is_sp2] <- ifelse(
    orthologs$Species2[is_sp2] %in% names(map),
    map[orthologs$Species2[is_sp2]], orthologs$Species2[is_sp2])
}
orthologs$sp1_name <- orthologs$sp2_name <- NULL
# Deduplicate after mapping (merged paralogs collapse to same pair)
orthologs <- unique(orthologs)

cat(nrow(orthologs), "ortholog pairs across",
    length(ortho_pairs), "species pairs\n")
```

## 2. Pairwise comparison and HOG-level testing

For each of the C(8,2) = 28 species pairs, we run two analyses:

1. **Neighborhood comparison** (`compare_neighborhoods()`): hypergeometric
   test for overlap of co-expression neighborhoods, producing per-pair
   p-values and fold-enrichment effect sizes.
2. **HOG-level permutation test** (`permutation_hog_test()`): gene-identity
   permutation within each HOG. This is necessary because pair-level tests
   within a HOG are correlated (shared neighborhoods), making Fisher's
   method or Storey q-values anti-conservative. The permutation p-value is
   exact regardless of the dependency structure.

With `use_torch = TRUE`, the fold-enrichment matrix is precomputed on
GPU, reducing each permutation to fast table lookups.

The HOG-level q-values determine which edges are co-expressologs
(conserved co-expression relationships):

```{r comparison}
# Helper: build edges from raw comparison + HOG-level q-values
build_perm_edges <- function(comp, hog_res, sp_a, sp_b, alpha = 0.05) {
  hog_q <- hog_res$q.value[match(comp$hog, hog_res$hog)]
  data.frame(
    gene1 = comp$Species1,
    gene2 = comp$Species2,
    species1 = sp_a,
    species2 = sp_b,
    hog = comp$hog,
    q.value = hog_q,
    effect_size = sqrt(comp$Species1.effect.size * comp$Species2.effect.size),
    type = ifelse(!is.na(hog_q) & hog_q < alpha, "conserved", "ns")
  )
}

# Run all 28 pairs
pair_edges <- list()
hog_results <- list()
idx <- 0L

for (pair in utils::combn(species, 2, simplify = FALSE)) {
  sp_a <- pair[1]; sp_b <- pair[2]

  comp <- compare_neighborhoods(networks[[sp_a]], networks[[sp_b]],
                                 orthologs)
  if (nrow(comp) == 0) next

  hog_res <- permutation_hog_test(
    networks[[sp_a]], networks[[sp_b]], comp,
    n_cores = min(4L, parallel::detectCores(logical = FALSE)),
    use_torch = TRUE
  )

  idx <- idx + 1L
  pair_edges[[idx]] <- build_perm_edges(comp, hog_res, sp_a, sp_b)
  hog_res$pair <- paste(sp_a, sp_b, sep = "-")
  hog_results[[paste(sp_a, sp_b, sep = "-")]] <- hog_res
}

edges <- do.call(rbind, pair_edges)
all_hog <- do.call(rbind, hog_results)

cat(nrow(edges), "edges:", sum(edges$type == "conserved"), "conserved,",
    sum(edges$type == "ns"), "non-significant\n")
cat(nrow(all_hog), "HOG tests,",
    sum(all_hog$q.value < 0.05), "significant (q < 0.05)\n")
```

For quick exploration without GPU, `run_pairwise_comparisons()` provides
a faster analytical path using pair-level Storey q-values (less
statistically rigorous for multi-copy HOGs):

```{r analytical-edges}
edges_analytical <- run_pairwise_comparisons(networks, orthologs)
cat("Analytical:", sum(edges_analytical$type == "conserved"), "conserved edges\n")
cat("Permutation:", sum(edges$type == "conserved"), "conserved edges\n")
```

## 4. Module-level analysis

Detect co-expression modules with Leiden community detection, then
compare across species:

```{r modules}
mod_bdis <- detect_modules(networks$BDIS, method = "leiden", resolution = 1.0)
mod_bsyl <- detect_modules(networks$BSYL, method = "leiden", resolution = 1.0)

mod_comp <- compare_modules(mod_bdis, mod_bsyl, orthologs,
                             method = "hypergeometric")
classes <- classify_modules(mod_comp)

show_table(as.data.frame(table(classes$classification)),
           caption = "Module classification (BDIS vs BSYL)")
```

### Hub genes and trait conservation

Identify hub genes within each species' modules, then classify each HOG
by whether it acts as a hub across trait groups. **Weighted degree** (the
default) is the recommended primary centrality for co-expression networks
-- it directly measures total co-expression strength to module neighbours,
which is the standard definition of a hub gene. All three centrality
measures (degree, betweenness, eigenvector) are reported regardless of
which is chosen as primary:

```{r module-hubs}
# Identify hubs (top 10% per module by weighted degree)
hubs_bdis <- identify_module_hubs(mod_bdis, networks$BDIS, orthologs,
                                   comparison = summary$results)
hubs_bsyl <- identify_module_hubs(mod_bsyl, networks$BSYL, orthologs,
                                   comparison = summary$results)

show_table(hubs_bdis[hubs_bdis$is_hub, ],
           caption = "Hub genes in BDIS modules")
```

```{r hub-conservation}
# Classify hub conservation across traits
trait <- c(BDIS = "annual", BSYL = "perennial")

hub_class <- classify_hub_conservation(
  list(BDIS = hubs_bdis, BSYL = hubs_bsyl), trait,
  module_comparisons = list("BDIS.BSYL" = mod_comp)
)

show_table(hub_class[hub_class$classification != "non_hub", ],
           caption = "Hub conservation (BDIS vs BSYL)")
```

## 5. Edge robustness: density sweep

**Network density** controls how many edges are retained in the
co-expression network. At 3% density, only the top 3% of all
pairwise MR values pass the threshold -- these are the strongest
co-expression relationships.

We sweep over 11 density levels centered at the 3% baseline,
testing both directions:

- **Tighter** (below 3%): does the signal survive stricter
  thresholds? Edges that persist are strong, not marginal.
- **Looser** (above 3%): is the signal specific to this density,
  or stable across a range? Cliques that only appear at very loose
  thresholds are likely noise from extra edges.

The 3% choice is somewhat arbitrary. Centering the sweep around it
reveals sensitivity to that choice -- a truly robust co-expressolog
should be present across a range of densities, not just at one
particular cutoff.

The permutation pipeline is re-run at each level because network
neighborhoods change with the threshold, altering the HOG-level
q-values:

```{r density-sweep}
# 11 levels: 5 looser + baseline + 5 tighter
multipliers <- seq(0.95, 1.05, by = 0.01)
sweep_results <- list()

for (m in multipliers) {
  tight_nets <- lapply(networks, function(net) {
    list(network = net$network, threshold = net$threshold * m)
  })
  names(tight_nets) <- species

  # Effective density at this threshold
  eff_density <- mean(vapply(species, function(sp) {
    vals <- networks[[sp]]$network[upper.tri(networks[[sp]]$network)]
    mean(vals >= networks[[sp]]$threshold * m)
  }, numeric(1)))

  # Re-run comparisons + permutation at stricter threshold
  tight_pair_edges <- list()
  tidx <- 0L
  for (pair in utils::combn(species, 2, simplify = FALSE)) {
    sp_a <- pair[1]; sp_b <- pair[2]
    comp <- tryCatch(
      compare_neighborhoods(tight_nets[[sp_a]], tight_nets[[sp_b]], orthologs),
      error = function(e) NULL)
    if (is.null(comp) || nrow(comp) == 0) next
    hog_res <- permutation_hog_test(tight_nets[[sp_a]], tight_nets[[sp_b]],
                                     comp,
                                     n_cores = min(4L, parallel::detectCores(logical = FALSE)),
                                     use_torch = TRUE)
    tidx <- tidx + 1L
    tight_pair_edges[[tidx]] <- build_perm_edges(comp, hog_res, sp_a, sp_b)
  }
  tight_edges <- if (tidx > 0) do.call(rbind, tight_pair_edges[seq_len(tidx)])
                 else edges[0, ]

  n_conserved <- sum(tight_edges$type == "conserved")
  sweep_results[[as.character(m)]] <- list(
    edges = tight_edges, eff_density = eff_density,
    n_conserved = n_conserved)
}

# Edge survival across density levels
sweep_df <- data.frame(
  Multiplier = paste0(names(sweep_results), "x"),
  Eff_density = sprintf("%.2f%%", vapply(sweep_results,
    function(r) r$eff_density * 100, numeric(1))),
  Conserved_edges = vapply(sweep_results, function(r) r$n_conserved, numeric(1)),
  Baseline = ifelse(names(sweep_results) == "1", "***", "")
)
show_table(sweep_df, caption = "Conserved edges across density levels")
```

The edge count may not decrease monotonically because the permutation
test operates at the HOG level: tighter neighborhoods can sharpen the
signal for some HOGs while eliminating it for others. Local fluctuations
reflect the discrete nature of the permutation q-values. The
interpretation is asymmetric: surviving tighter thresholds indicates
strong signal, while surviving looser thresholds indicates specificity
(not merely noise from additional edges).

**Analytical alternative: `clique_persistence()`.**
The density sweep above is rigorous but expensive — it re-runs the
full permutation pipeline at each threshold level. For a quick estimate,
`clique_persistence()` computes the ratio of each co-expressolog edge's
MR value to the species' density threshold. A persistence of 1.5 means
the weakest supporting edge is 50% above threshold and would survive a
1.5x tightening. This is a principled analytical proxy: it correctly
identifies the weakest link in each clique's co-expressolog support.
However, it cannot detect that tighter neighborhoods may change which
HOGs reach significance in the permutation test, or that new cliques
may form from HOGs that become significant at stricter thresholds.
Use persistence for screening or when GPU is unavailable; use the full
sweep for publication-quality robustness assessment.

## 6. Clique detection and phylogenetic stability

Using edges from the baseline density level, we find cliques and
test their phylogenetic stability via leave-k-out jackknife. The
jackknife removes k species at a time and checks whether each
trait-exclusive clique is preserved:

```{r cliques-stability}
# Use the baseline (1.0x) edges for clique detection and stability
robust_edges <- sweep_results[["1"]]$edges
cat("Using", sum(robust_edges$type == "conserved"), "conserved edges at",
    sprintf("%.2f%%", sweep_results[["1"]]$eff_density * 100), "density\n\n")

# Find cliques
cliques <- find_cliques(robust_edges, species, min_species = 2L)
cat(nrow(cliques), "cliques\n")

show_table(as.data.frame(table(n_species = cliques$n_species)),
           pageLength = 10, caption = "Cliques by number of species")
```

```{r stability}
# Jackknife stability (all k from 1 to N-2)
max_k <- length(species) - 2L
stab <- clique_stability(robust_edges, species, trait,
                          all_species = species,
                          full_cliques = cliques,
                          max_k = max_k)
cat("Trait-exclusive cliques:", length(stab$stability_class), "\n")

stab_df <- as.data.frame(table(stability_class = factor(
  stab$stability_class, levels = 0:max_k)))
show_table(stab_df, pageLength = 10, caption = "Stability class distribution")
```

## 7. HOG classification with robustness ranking

Classify each HOG by its conservation pattern, then rank within each
category by robustness. The waterfall assigns each HOG to exactly one
class: complete > partial > differentiated > trait_specific > unclassified.

```{r classify}
result <- classify_cliques(robust_edges, species, trait,
                            min_species = 2L,
                            stability = stab)

show_table(as.data.frame(table(classification = result$classification)),
           pageLength = 5, caption = "HOG classification summary")
```

- **complete**: all 8 species form a clique -- deeply conserved across
  all Pooideae
- **partial**: clique spans both annual and perennial species, but not
  all 8
- **differentiated**: annual species have a clique AND perennial species
  have a clique, but no cross-group conservation -- lineage-specific
  co-expression programs
- **trait_specific**: only annuals or only perennials have a clique
- **unclassified**: no conservation signal detected

The robustness ranking comes from combining density sweep survival
with stability class. HOGs classified at a stricter density level
with high stability class are the strongest findings:

```{r ranking}
# Count how many density levels each HOG's edges survive
hog_edge_survival <- cbind(vapply(names(sweep_results), function(m) {
  e <- sweep_results[[m]]$edges
  conserved_hogs <- unique(e$hog[e$type == "conserved"])
  result$hog %in% conserved_hogs
}, logical(nrow(result))))
result$n_densities_survived <- rowSums(hog_edge_survival)

# Annotate trait composition from the baseline cliques
# For each HOG, count how many annual/perennial species are in its best clique
result$n_annual <- NA_integer_
result$n_perennial <- NA_integer_
for (i in seq_len(nrow(result))) {
  hog_cliques <- cliques[cliques$hog == result$hog[i], , drop = FALSE]
  if (nrow(hog_cliques) == 0) next
  idx <- which.min(hog_cliques$mean_q)
  if (length(idx) == 0) next
  best <- hog_cliques[idx, ]
  present_sp <- species[!is.na(best[, species])]
  result$n_annual[i] <- sum(trait[present_sp] == "annual")
  result$n_perennial[i] <- sum(trait[present_sp] == "perennial")
}

# Show ranked HOGs per category
show_cols <- c("hog", "n_species", "n_annual", "n_perennial",
               "best_mean_q", "stability_class", "n_densities_survived")
show_cols <- intersect(show_cols, names(result))
```

Note that `stability_class` is NA for partial and complete cliques --
these span both trait groups and are therefore not trait-exclusive. The
jackknife stability test asks "does removing species break trait
exclusivity?", which only applies to trait-exclusive cliques
(differentiated and trait_specific). For cross-trait cliques, the
`n_densities_survived` column is the relevant robustness measure.

### {.tabset}

```{r partial, results='asis'}
sub <- result[result$classification == "partial", show_cols]
if (nrow(sub) > 0) {
  sub <- sub[order(-sub$n_densities_survived,
                   if ("best_mean_q" %in% names(sub)) sub$best_mean_q else 0), ]
}
```

#### Partial

```{r partial-table}
show_table(sub, pageLength = 10, caption = "Partial HOGs (cross-trait cliques)")
```

```{r trait-specific, results='asis'}
sub_ts <- result[result$classification == "trait_specific", show_cols]
if (nrow(sub_ts) > 0) {
  sub_ts <- sub_ts[order(-sub_ts$n_densities_survived,
                         if ("best_mean_q" %in% names(sub_ts)) sub_ts$best_mean_q else 0), ]
}
```

#### Trait-specific

```{r trait-specific-table}
show_table(sub_ts, pageLength = 10,
           caption = "Trait-specific HOGs (one lineage only)")
```

```{r differentiated, results='asis'}
sub_diff <- result[result$classification == "differentiated", show_cols]
if (nrow(sub_diff) > 0) {
  sub_diff <- sub_diff[order(-sub_diff$n_densities_survived,
                             if ("best_mean_q" %in% names(sub_diff)) sub_diff$best_mean_q else 0), ]
}
```

#### Differentiated

```{r differentiated-table}
show_table(sub_diff, pageLength = 10,
           caption = "Differentiated HOGs (within-group conservation, cross-group divergence)")
```

```{r complete, results='asis'}
sub_comp <- result[result$classification == "complete", show_cols]
if (nrow(sub_comp) > 0) {
  sub_comp <- sub_comp[order(-sub_comp$n_densities_survived,
                             if ("best_mean_q" %in% names(sub_comp)) sub_comp$best_mean_q else 0), ]
}
```

#### Complete

```{r complete-table}
show_table(sub_comp, pageLength = 10,
           caption = "Complete HOGs (all 8 species)")
```

###

## Session info

```{r sessioninfo}
sessionInfo()
```
