---
title: "Comparative co-expression analysis with rcomplex"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparative co-expression analysis with rcomplex}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette demonstrates a complete rcomplex analysis using expression
data from eight Pooideae grass species spanning four phylogenetic pairs,
each containing one annual and one perennial species:

| Species | Code | Life cycle | Pair |
|---------|------|------------|------|
| *Brachypodium distachyon* | BDIS | Annual | Brachypodium |
| *Brachypodium sylvaticum* | BSYL | Perennial | Brachypodium |
| *Hordeum vulgare* | HVUL | Annual | Hordeum |
| *Hordeum jubatum* | HJUB | Perennial | Hordeum |
| *Briza maxima* | BMAX | Annual | Briza |
| *Briza media* (crossmapped to BMAX) | BMED | Perennial | Briza |
| *Vulpia bromoides* | VBRO | Annual | Festuca/Vulpia |
| *Festuca pratensis* | FPRA | Perennial | Festuca/Vulpia |

The dataset is a 500-HOG subset of VST-normalized RNA-seq from the
EVOTREE project, stored as SummarizedExperiment objects.

## 1. Data preparation

```{r data}
library(rcomplex)
library(SummarizedExperiment)

se_list <- readRDS(system.file("extdata", "pooideae_vignette.rds",
                                package = "rcomplex"))
species <- names(se_list)

# Life cycle trait
trait <- c(BDIS = "annual", BSYL = "perennial",
           HVUL = "annual", HJUB = "perennial",
           BMAX = "annual", BMED = "perennial",
           VBRO = "annual", FPRA = "perennial")
```

Build co-expression networks for each species using Mutual Rank (MR)
normalization at 3% density:

```{r networks}
networks <- lapply(species, function(sp) {
  compute_network(se_list[[sp]], norm_method = "MR", density = 0.03)
})
names(networks) <- species
```

Extract ortholog pairs from HOG annotations in rowData:

```{r orthologs}
ortho_pairs <- utils::combn(species, 2, simplify = FALSE)
ortho_list <- lapply(ortho_pairs, function(pair) {
  extract_orthologs(se_list[[pair[1]]], se_list[[pair[2]]])
})
orthologs <- do.call(rbind, ortho_list)
cat(nrow(orthologs), "ortholog pairs across",
    length(ortho_pairs), "species pairs\n")
```

## 2. Pairwise comparison

`run_pairwise_comparisons()` runs the full comparison pipeline for all
C(8,2) = 28 species pairs: hypergeometric neighborhood test, Storey
q-values, and edge classification:

```{r comparison}
edges <- run_pairwise_comparisons(networks, orthologs)
cat(nrow(edges), "edges:", sum(edges$type == "conserved"), "conserved,",
    sum(edges$type == "ns"), "non-significant\n")
```

## 3. Gene/HOG-level analysis

For a single species pair, `summarize_comparison()` provides detailed
statistics. The `sp1`/`sp2` arguments produce edge-format output directly:

```{r gene-level}
comp_bdis_bsyl <- compare_neighborhoods(
  networks$BDIS, networks$BSYL, orthologs)
summary_bb <- summarize_comparison(comp_bdis_bsyl,
                                    sp1 = "BDIS", sp2 = "BSYL")
cat("Reciprocally conserved pairs:", summary_bb$summary$gene_pairs$reciprocal,
    "of", summary_bb$summary$gene_pairs$total, "\n")
```

## 4. Module-level analysis

Detect co-expression modules with Leiden community detection, then
compare across species:

```{r modules}
mod_bdis <- detect_modules(networks$BDIS, method = "leiden", resolution = 1.0)
mod_bsyl <- detect_modules(networks$BSYL, method = "leiden", resolution = 1.0)

mod_comp <- compare_modules(mod_bdis, mod_bsyl, orthologs,
                             method = "hypergeometric")
classes <- classify_modules(mod_comp)
print(table(classes$classification))
```

## 5. Clique detection

Find fully connected subsets of ortholog groups across species:

```{r cliques}
# All cliques (at least 2 species)
cliques <- find_cliques(edges, species, min_species = 2L)
cat(nrow(cliques), "cliques\n")
print(table(cliques$n_species))
```

Analytical persistence scores measure how far above the detection
threshold each clique's supporting edges are:

```{r persistence}
cliques_p <- clique_persistence(cliques, species, networks, edges)
cat("Median persistence:", round(median(cliques_p$persistence, na.rm = TRUE), 2), "\n")
```

## 6. Combined robustness analysis

The strongest findings are cliques robust on two independent axes:

- **Threshold robustness** -- survives at stricter density thresholds
- **Phylogenetic robustness** -- survives species removal (jackknife)

We sweep over threshold multipliers and run leave-k-out stability at
each level:

```{r robustness}
multipliers <- c(1.5, 2, 3)
robustness <- list()

for (m in multipliers) {
  # Re-threshold networks (shallow copy, no matrix duplication)
  tight_nets <- lapply(networks, function(net) {
    list(network = net$network, threshold = net$threshold * m)
  })
  names(tight_nets) <- species

  # Re-run comparisons at stricter threshold
  tight_edges <- run_pairwise_comparisons(tight_nets, orthologs)

  # Find cliques at stricter threshold
  tight_cliques <- find_cliques(tight_edges, species, min_species = 2L)

  # Jackknife stability at this threshold level
  if (nrow(tight_cliques) > 0) {
    stab <- clique_stability(tight_edges, species, trait,
                              all_species = species,
                              full_cliques = tight_cliques,
                              max_k = 1L)
    robustness[[as.character(m)]] <- list(
      cliques = tight_cliques,
      n_cliques = nrow(tight_cliques),
      n_stable = sum(stab$stability_class >= 1),
      stability = stab
    )
  } else {
    robustness[[as.character(m)]] <- list(
      cliques = tight_cliques, n_cliques = 0L, n_stable = 0L)
  }
}

# Summary: cliques that survive AND are phylogenetically stable
cat("Multiplier | Cliques | Stable (k=1)\n")
cat("-----------|---------|-------------\n")
for (m in names(robustness)) {
  r <- robustness[[m]]
  cat(sprintf("%-10s | %7d | %12d\n",
              paste0(m, "x"), r$n_cliques, r$n_stable))
}
```

## 7. HOG classification

Classify each HOG by its conservation pattern using the waterfall:
complete > partial > differentiated > trait_specific > unclassified.

```{r classify}
# Use stability from the 1.5x threshold level
stab_input <- if (!is.null(robustness[["1.5"]]$stability)) {
  robustness[["1.5"]]$stability
} else NULL

result <- classify_cliques(edges, species, trait,
                            min_species = 2L,
                            stability = stab_input)
print(table(result$classification))
```

- **complete**: all 8 species form a clique -- deeply conserved across
  all Pooideae
- **partial**: clique spans both annual and perennial species, but not
  all 8
- **differentiated**: annual species have a clique AND perennial species
  have a clique, but no cross-group conservation -- lineage-specific
  co-expression programs
- **trait_specific**: only annuals or only perennials have a clique
- **unclassified**: no conservation signal detected

## Session info

```{r sessioninfo}
sessionInfo()
```
