---
title: "Comparative co-expression analysis with rcomplex"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparative co-expression analysis with rcomplex}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  eval = requireNamespace("SummarizedExperiment", quietly = TRUE)
)
```

This vignette demonstrates a complete rcomplex analysis using expression
data from eight Pooideae grass species spanning four phylogenetic pairs,
each containing one annual and one perennial species:

| Species | Code | Life cycle | Pair |
|---------|------|------------|------|
| *Brachypodium distachyon* | BDIS | Annual | Brachypodium |
| *Brachypodium sylvaticum* | BSYL | Perennial | Brachypodium |
| *Hordeum vulgare* | HVUL | Annual | Hordeum |
| *Hordeum jubatum* | HJUB | Perennial | Hordeum |
| *Briza maxima* | BMAX | Annual | Briza |
| *Briza media* (crossmapped to BMAX) | BMED | Perennial | Briza |
| *Vulpia bromoides* | VBRO | Annual | Festuca/Vulpia |
| *Festuca pratensis* | FPRA | Perennial | Festuca/Vulpia |

The dataset is a 500-HOG subset of VST-normalized RNA-seq from the
"Live or let die?" project, stored as SummarizedExperiment objects.

## 1. Data preparation

```{r data}
library(rcomplex)
library(SummarizedExperiment)

se_list <- readRDS(system.file("extdata", "pooideae_vignette.rds",
                                package = "rcomplex"))
species <- names(se_list)

# Life cycle trait
trait <- c(BDIS = "annual", BSYL = "perennial",
           HVUL = "annual", HJUB = "perennial",
           BMAX = "annual", BMED = "perennial",
           VBRO = "annual", FPRA = "perennial")
```

Build co-expression networks for each species using Mutual Rank (MR)
normalization at 3% density:

```{r networks}
networks <- lapply(species, function(sp) {
  compute_network(se_list[[sp]], norm_method = "MR", density = 0.03)
})
names(networks) <- species
```

Extract ortholog pairs from HOG annotations in rowData:

```{r orthologs}
ortho_pairs <- utils::combn(species, 2, simplify = FALSE)
ortho_list <- lapply(ortho_pairs, function(pair) {
  extract_orthologs(se_list[[pair[1]]], se_list[[pair[2]]])
})
orthologs <- do.call(rbind, ortho_list)
cat(nrow(orthologs), "ortholog pairs across",
    length(ortho_pairs), "species pairs\n")
```

## 2. Pairwise comparison

`run_pairwise_comparisons()` runs the full comparison pipeline for all
C(8,2) = 28 species pairs: hypergeometric neighborhood test, Storey
q-values, and edge classification:

```{r comparison}
edges <- run_pairwise_comparisons(networks, orthologs)
cat(nrow(edges), "edges:", sum(edges$type == "conserved"), "conserved,",
    sum(edges$type == "ns"), "non-significant\n")
```

## 3. Gene/HOG-level analysis

For a single species pair, `summarize_comparison()` provides detailed
statistics. The `sp1`/`sp2` arguments produce edge-format output directly:

```{r gene-level}
comp_bdis_bsyl <- compare_neighborhoods(
  networks$BDIS, networks$BSYL, orthologs)
summary_bb <- summarize_comparison(comp_bdis_bsyl,
                                    sp1 = "BDIS", sp2 = "BSYL")
cat("Reciprocally conserved pairs:", summary_bb$summary$gene_pairs$reciprocal,
    "of", summary_bb$summary$gene_pairs$total, "\n")
```

### HOG-level permutation test (GPU-accelerated)

The pair-level hypergeometric test treats each gene pair independently,
but genes within the same HOG share network neighborhoods. Fisher's
method for combining these correlated p-values is anti-conservative.
`permutation_hog_test()` instead permutes gene identities within each
HOG, producing exact p-values regardless of the dependency structure.

With `use_torch = TRUE`, the fold-enrichment matrix is precomputed on
GPU via matrix multiplication, reducing each permutation to fast table
lookups. We run this for all 28 species pairs:

```{r permutation, eval = requireNamespace("torch", quietly = TRUE)}
hog_results <- list()
for (pair in utils::combn(species, 2, simplify = FALSE)) {
  sp_a <- pair[1]; sp_b <- pair[2]
  pair_name <- paste(sp_a, sp_b, sep = "-")

  comp <- compare_neighborhoods(networks[[sp_a]], networks[[sp_b]],
                                 orthologs)
  if (nrow(comp) == 0) next

  hog_res <- permutation_hog_test(
    networks[[sp_a]], networks[[sp_b]], comp,
    n_cores = 4L, use_torch = TRUE
  )
  hog_res$pair <- pair_name
  hog_results[[pair_name]] <- hog_res
}

all_hog <- do.call(rbind, hog_results)
cat(nrow(all_hog), "HOG tests across", length(hog_results), "pairs\n")
cat(sum(all_hog$q.value < 0.05), "significant HOGs (q < 0.05)\n")
```

## 4. Module-level analysis

Detect co-expression modules with Leiden community detection, then
compare across species:

```{r modules}
mod_bdis <- detect_modules(networks$BDIS, method = "leiden", resolution = 1.0)
mod_bsyl <- detect_modules(networks$BSYL, method = "leiden", resolution = 1.0)

mod_comp <- compare_modules(mod_bdis, mod_bsyl, orthologs,
                             method = "hypergeometric")
classes <- classify_modules(mod_comp)
print(table(classes$classification))
```

## 5. Clique detection

Find fully connected subsets of ortholog groups across species:

```{r cliques}
# All cliques (at least 2 species)
cliques <- find_cliques(edges, species, min_species = 2L)
cat(nrow(cliques), "cliques\n")
print(table(cliques$n_species))
```

Analytical persistence scores measure how far above the detection
threshold each clique's supporting edges are:

```{r persistence}
cliques_p <- clique_persistence(cliques, species, networks, edges)
cat("Median persistence:", round(median(cliques_p$persistence, na.rm = TRUE), 2), "\n")
```

## 6. Combined robustness analysis

The strongest findings are cliques robust on two independent axes:

- **Threshold robustness** -- the clique survives at stricter density
  thresholds (the co-expression signal is strong, not marginal)
- **Phylogenetic robustness** -- the clique survives species removal
  (the conservation pattern is general, not taxon-dependent)

**Network density** controls how many edges are retained in the
co-expression network. At 3% density, only the top 3% of all
pairwise MR values pass the threshold -- these are the strongest
co-expression relationships. A gene pair that barely passes at 3%
might not pass at 2.5%, while a truly co-expressed pair will survive
much stricter cutoffs.

We test robustness by sweeping over tighter density percentiles and
running leave-k-out jackknife stability at each level:

```{r robustness}
# The baseline density is 3%. Each multiplier tightens the MR threshold,
# which corresponds to a lower effective density percentile.
multipliers <- c(1.01, 1.02, 1.03, 1.05)
robustness <- list()

for (m in multipliers) {

  # Re-threshold: only $network and $threshold are needed downstream
  tight_nets <- lapply(networks, function(net) {
    list(network = net$network, threshold = net$threshold * m)
  })
  names(tight_nets) <- species

  # Compute effective density at this threshold (fraction of edges surviving)
  eff_densities <- vapply(species, function(sp) {
    mat <- networks[[sp]]$network
    vals <- mat[upper.tri(mat)]
    mean(vals >= networks[[sp]]$threshold * m)
  }, numeric(1))

  # Re-run comparisons at stricter threshold
  tight_edges <- run_pairwise_comparisons(tight_nets, orthologs)

  # Find cliques at stricter threshold
  tight_cliques <- find_cliques(tight_edges, species, min_species = 2L)

  # Jackknife stability: test all k from 1 to max (N - 2 = 6)
  if (nrow(tight_cliques) > 0) {
    max_k <- length(species) - 2L
    stab <- clique_stability(tight_edges, species, trait,
                              all_species = species,
                              full_cliques = tight_cliques,
                              max_k = max_k)
    robustness[[as.character(m)]] <- list(
      cliques = tight_cliques,
      n_cliques = nrow(tight_cliques),
      n_exclusive = length(stab$stability_class),
      stability_class = stab$stability_class,
      mean_density = mean(eff_densities),
      stability = stab
    )
  } else {
    robustness[[as.character(m)]] <- list(
      cliques = tight_cliques, n_cliques = 0L, n_exclusive = 0L,
      stability_class = integer(0),
      mean_density = mean(eff_densities))
  }
}

# Summary: cliques surviving at each density level with stability class
# stability_class = highest k at which ALL species-removal subsets preserve
# the clique's trait exclusivity. k=6 means the clique survives removal of
# any 6 of 8 species (maximally robust).
max_k <- length(species) - 2L
header <- paste0("Density    | Cliques | Exclusive | ",
                 paste(sprintf("k=%d", 0:max_k), collapse = " | "))
cat(header, "\n")
cat(paste(rep("-", nchar(header)), collapse = ""), "\n")
for (m in names(robustness)) {
  r <- robustness[[m]]
  sc_tab <- table(factor(r$stability_class, levels = 0:max_k))
  sc_str <- paste(sprintf("%3d", sc_tab), collapse = " | ")
  cat(sprintf("%-10s | %7d | %9d | %s\n",
              sprintf("%.2f%%", r$mean_density * 100),
              r$n_cliques, r$n_exclusive, sc_str))
}
```

The stability class distribution is strikingly bimodal: cliques are either
unstable (class 0) or maximally stable (class `r length(species) - 2L`).
This makes biological sense -- with 4 species per trait group, a
trait-exclusive clique either depends on a specific species combination
(fragile) or reflects a genuinely conserved pattern that persists regardless
of which species are removed (robust).

Cliques that survive at tighter density AND have maximal stability class
represent the strongest conservation signals in the dataset.

## 7. HOG classification

Classify each HOG by its conservation pattern using the waterfall:
complete > partial > differentiated > trait_specific > unclassified.

```{r classify}
# Use stability from the 1.02x level (NULL if no cliques survived)
stab_input <- robustness[["1.02"]]$stability

result <- classify_cliques(edges, species, trait,
                            min_species = 2L,
                            stability = stab_input)
print(table(result$classification))
```

- **complete**: all 8 species form a clique -- deeply conserved across
  all Pooideae
- **partial**: clique spans both annual and perennial species, but not
  all 8
- **differentiated**: annual species have a clique AND perennial species
  have a clique, but no cross-group conservation -- lineage-specific
  co-expression programs
- **trait_specific**: only annuals or only perennials have a clique
- **unclassified**: no conservation signal detected

## Session info

```{r sessioninfo}
sessionInfo()
```
