---
title: "Comparative co-expression analysis with rcomplex"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparative co-expression analysis with rcomplex}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  eval = requireNamespace("SummarizedExperiment", quietly = TRUE) &&
         requireNamespace("torch", quietly = TRUE)
)
```

This vignette demonstrates a complete rcomplex analysis using expression
data from eight Pooideae grass species spanning four phylogenetic pairs,
each containing one annual and one perennial species:

| Species | Code | Life cycle | Pair |
|---------|------|------------|------|
| *Brachypodium distachyon* | BDIS | Annual | Brachypodium |
| *Brachypodium sylvaticum* | BSYL | Perennial | Brachypodium |
| *Hordeum vulgare* | HVUL | Annual | Hordeum |
| *Hordeum jubatum* | HJUB | Perennial | Hordeum |
| *Briza maxima* | BMAX | Annual | Briza |
| *Briza media* (crossmapped to BMAX) | BMED | Perennial | Briza |
| *Vulpia bromoides* | VBRO | Annual | Festuca/Vulpia |
| *Festuca pratensis* | FPRA | Perennial | Festuca/Vulpia |

The dataset is a 500-HOG subset of VST-normalized RNA-seq from the
"Live or let die?" project, stored as SummarizedExperiment objects.

## 1. Data preparation

```{r data, message = FALSE}
library(rcomplex)
library(SummarizedExperiment)

se_list <- readRDS(system.file("extdata", "pooideae_vignette.rds",
                                package = "rcomplex"))
species <- names(se_list)

# Life cycle trait
trait <- c(BDIS = "annual", BSYL = "perennial",
           HVUL = "annual", HJUB = "perennial",
           BMAX = "annual", BMED = "perennial",
           VBRO = "annual", FPRA = "perennial")
```

Build co-expression networks for each species using Mutual Rank (MR)
normalization at 3% density:

```{r networks}
networks <- lapply(species, function(sp) {
  compute_network(se_list[[sp]], norm_method = "MR", density = 0.03)
})
names(networks) <- species
```

Extract ortholog pairs from HOG annotations in rowData:

```{r orthologs}
ortho_pairs <- utils::combn(species, 2, simplify = FALSE)
ortho_list <- lapply(ortho_pairs, function(pair) {
  extract_orthologs(se_list[[pair[1]]], se_list[[pair[2]]])
})
orthologs <- do.call(rbind, ortho_list)
cat(nrow(orthologs), "ortholog pairs across",
    length(ortho_pairs), "species pairs\n")
```

## 2. Pairwise comparison and HOG-level testing

For each of the C(8,2) = 28 species pairs, we run two analyses:

1. **Neighborhood comparison** (`compare_neighborhoods()`): hypergeometric
   test for overlap of co-expression neighborhoods, producing per-pair
   p-values and fold-enrichment effect sizes.
2. **HOG-level permutation test** (`permutation_hog_test()`): gene-identity
   permutation within each HOG. This is necessary because pair-level tests
   within a HOG are correlated (shared neighborhoods), making Fisher's
   method or Storey q-values anti-conservative. The permutation p-value is
   exact regardless of the dependency structure.

With `use_torch = TRUE`, the fold-enrichment matrix is precomputed on
GPU, reducing each permutation to fast table lookups.

The HOG-level q-values determine which edges are co-expressologs
(conserved co-expression relationships):

```{r comparison}
# Helper: build edges from raw comparison + HOG-level q-values
build_perm_edges <- function(comp, hog_res, sp_a, sp_b, alpha = 0.05) {
  hog_q <- hog_res$q.value[match(comp$hog, hog_res$hog)]
  data.frame(
    gene1 = comp$Species1,
    gene2 = comp$Species2,
    species1 = sp_a,
    species2 = sp_b,
    hog = comp$hog,
    q.value = hog_q,
    effect_size = sqrt(comp$Species1.effect.size * comp$Species2.effect.size),
    type = ifelse(!is.na(hog_q) & hog_q < alpha, "conserved", "ns")
  )
}

# Run all 28 pairs
pair_edges <- list()
hog_results <- list()
idx <- 0L

for (pair in utils::combn(species, 2, simplify = FALSE)) {
  sp_a <- pair[1]; sp_b <- pair[2]

  comp <- compare_neighborhoods(networks[[sp_a]], networks[[sp_b]],
                                 orthologs)
  if (nrow(comp) == 0) next

  hog_res <- permutation_hog_test(
    networks[[sp_a]], networks[[sp_b]], comp,
    n_cores = min(4L, parallel::detectCores(logical = FALSE)),
    use_torch = TRUE
  )

  idx <- idx + 1L
  pair_edges[[idx]] <- build_perm_edges(comp, hog_res, sp_a, sp_b)
  hog_res$pair <- paste(sp_a, sp_b, sep = "-")
  hog_results[[paste(sp_a, sp_b, sep = "-")]] <- hog_res
}

edges <- do.call(rbind, pair_edges)
all_hog <- do.call(rbind, hog_results)

cat(nrow(edges), "edges:", sum(edges$type == "conserved"), "conserved,",
    sum(edges$type == "ns"), "non-significant\n")
cat(nrow(all_hog), "HOG tests,",
    sum(all_hog$q.value < 0.05), "significant (q < 0.05)\n")
```

For quick exploration without GPU, `run_pairwise_comparisons()` provides
a faster analytical path using pair-level Storey q-values (less
statistically rigorous for multi-copy HOGs):

```{r analytical-edges}
edges_analytical <- run_pairwise_comparisons(networks, orthologs)
cat("Analytical:", sum(edges_analytical$type == "conserved"), "conserved edges\n")
cat("Permutation:", sum(edges$type == "conserved"), "conserved edges\n")
```

## 4. Module-level analysis

Detect co-expression modules with Leiden community detection, then
compare across species:

```{r modules}
mod_bdis <- detect_modules(networks$BDIS, method = "leiden", resolution = 1.0)
mod_bsyl <- detect_modules(networks$BSYL, method = "leiden", resolution = 1.0)

mod_comp <- compare_modules(mod_bdis, mod_bsyl, orthologs,
                             method = "hypergeometric")
classes <- classify_modules(mod_comp)
print(table(classes$classification))
```

## 5. Edge robustness: density sweep

**Network density** controls how many edges are retained. At 3%
density, only the top 3% of pairwise MR values pass the threshold.
A co-expressolog edge that barely passes at 3% may be noise; one
that survives at 1% density is a strong signal.

We sweep over tighter density percentiles, re-running the full
permutation pipeline at each level (neighborhoods change with
threshold, so permutation q-values must be recomputed):

```{r density-sweep}
multipliers <- c(1.01, 1.02, 1.03, 1.05)
sweep_results <- list()

for (m in multipliers) {
  tight_nets <- lapply(networks, function(net) {
    list(network = net$network, threshold = net$threshold * m)
  })
  names(tight_nets) <- species

  # Effective density at this threshold
  eff_density <- mean(vapply(species, function(sp) {
    vals <- networks[[sp]]$network[upper.tri(networks[[sp]]$network)]
    mean(vals >= networks[[sp]]$threshold * m)
  }, numeric(1)))

  # Re-run comparisons + permutation at stricter threshold
  tight_pair_edges <- list()
  tidx <- 0L
  for (pair in utils::combn(species, 2, simplify = FALSE)) {
    sp_a <- pair[1]; sp_b <- pair[2]
    comp <- tryCatch(
      compare_neighborhoods(tight_nets[[sp_a]], tight_nets[[sp_b]], orthologs),
      error = function(e) NULL)
    if (is.null(comp) || nrow(comp) == 0) next
    hog_res <- permutation_hog_test(tight_nets[[sp_a]], tight_nets[[sp_b]],
                                     comp, n_cores = 4L, use_torch = TRUE)
    tidx <- tidx + 1L
    tight_pair_edges[[tidx]] <- build_perm_edges(comp, hog_res, sp_a, sp_b)
  }
  tight_edges <- if (tidx > 0) do.call(rbind, tight_pair_edges[seq_len(tidx)])
                 else edges[0, ]

  n_conserved <- sum(tight_edges$type == "conserved")
  sweep_results[[as.character(m)]] <- list(
    edges = tight_edges, eff_density = eff_density,
    n_conserved = n_conserved)
}

# Edge survival across density levels
cat("Density | Conserved edges\n")
cat("--------|----------------\n")
cat(sprintf("%-7s | %15d  (baseline)\n", "3.00%", sum(edges$type == "conserved")))
for (m in names(sweep_results)) {
  r <- sweep_results[[m]]
  cat(sprintf("%-7s | %15d\n", sprintf("%.2f%%", r$eff_density * 100), r$n_conserved))
}
```

## 6. Clique detection and phylogenetic stability

Using edges from a robust density level (1.02x), we find cliques and
test their phylogenetic stability via leave-k-out jackknife. The
jackknife removes k species at a time and checks whether each
trait-exclusive clique is preserved:

```{r cliques-stability}
# Pick a density level where signal is robust but not too sparse
robust_edges <- sweep_results[["1.02"]]$edges
cat("Using", sum(robust_edges$type == "conserved"), "conserved edges at",
    sprintf("%.2f%%", sweep_results[["1.02"]]$eff_density * 100), "density\n\n")

# Find cliques
cliques <- find_cliques(robust_edges, species, min_species = 2L)
cat(nrow(cliques), "cliques\n")
print(table(cliques$n_species))

# Jackknife stability (all k from 1 to N-2)
max_k <- length(species) - 2L
stab <- clique_stability(robust_edges, species, trait,
                          all_species = species,
                          full_cliques = cliques,
                          max_k = max_k)
cat("\nTrait-exclusive cliques:", length(stab$stability_class), "\n")
cat("Stability class distribution:\n")
print(table(factor(stab$stability_class, levels = 0:max_k)))
```

## 7. HOG classification with robustness ranking

Classify each HOG by its conservation pattern, then rank within each
category by robustness. The waterfall assigns each HOG to exactly one
class: complete > partial > differentiated > trait_specific > unclassified.

```{r classify}
result <- classify_cliques(robust_edges, species, trait,
                            min_species = 2L,
                            stability = stab)
print(table(result$classification))
```

- **complete**: all 8 species form a clique -- deeply conserved across
  all Pooideae
- **partial**: clique spans both annual and perennial species, but not
  all 8
- **differentiated**: annual species have a clique AND perennial species
  have a clique, but no cross-group conservation -- lineage-specific
  co-expression programs
- **trait_specific**: only annuals or only perennials have a clique
- **unclassified**: no conservation signal detected

The robustness ranking comes from combining density sweep survival
with stability class. HOGs classified at a stricter density level
with high stability class are the strongest findings:

```{r ranking}
# Count how many density levels each HOG's edges survive
hog_edge_survival <- vapply(names(sweep_results), function(m) {
  e <- sweep_results[[m]]$edges
  conserved_hogs <- unique(e$hog[e$type == "conserved"])
  result$hog %in% conserved_hogs
}, logical(nrow(result)))
result$n_densities_survived <- rowSums(hog_edge_survival)

# Show top-ranked HOGs per category
for (cls in c("complete", "partial", "differentiated", "trait_specific")) {
  sub <- result[result$classification == cls, ]
  if (nrow(sub) == 0) next
  sub <- sub[order(-sub$n_densities_survived, sub$best_mean_q), ]
  cat("\n", toupper(cls), "- top 5:\n")
  show_cols <- c("hog", "n_species", "best_mean_q", "stability_class",
                 "n_densities_survived")
  show_cols <- intersect(show_cols, names(sub))
  print(head(sub[, show_cols], 5), row.names = FALSE)
}
```

## Session info

```{r sessioninfo}
sessionInfo()
```
