---
title: "Comparative co-expression analysis with rcomplex"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparative co-expression analysis with rcomplex}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_se <- requireNamespace("SummarizedExperiment", quietly = TRUE)
has_torch <- requireNamespace("torch", quietly = TRUE) &&
             torch::torch_is_installed() &&
             tryCatch({ torch::torch_randn(1); TRUE }, error = function(e) FALSE)
has_tidy <- requireNamespace("purrr", quietly = TRUE) &&
            requireNamespace("tibble", quietly = TRUE)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      eval = has_se && has_torch && has_tidy)

# Helper for interactive tables (sortable by one or multiple columns)
show_table <- function(df, pageLength = 10, ...) {
  if (requireNamespace("DT", quietly = TRUE)) {
    DT::datatable(df, rownames = FALSE,
                  options = list(pageLength = pageLength, scrollX = TRUE,
                                 ordering = TRUE, orderMulti = TRUE),
                  ...)
  } else {
    knitr::kable(df)
  }
}

# Helper for ranked classification tables
show_classification <- function(df, class, cols, caption) {
  sub <- df |>
    dplyr::filter(classification == class) |>
    dplyr::select(dplyr::any_of(cols)) |>
    dplyr::arrange(dplyr::desc(n_species), dplyr::desc(n_densities_survived), best_mean_q)
  show_table(sub, pageLength = 10, caption = caption)
}
```

This vignette demonstrates a complete rcomplex analysis using expression
data from eight Pooideae grass species spanning four phylogenetic pairs,
each containing one annual and one perennial species:

| Species | Code | Life cycle | Pair |
|---------|------|------------|------|
| *Brachypodium distachyon* | BDIS | Annual | Brachypodium |
| *Brachypodium sylvaticum* | BSYL | Perennial | Brachypodium |
| *Hordeum vulgare* | HVUL | Annual | Hordeum |
| *Hordeum jubatum* | HJUB | Perennial | Hordeum |
| *Briza maxima* | BMAX | Annual | Briza |
| *Briza media* (crossmapped to BMAX) | BMED | Perennial | Briza |
| *Vulpia bromoides* | VBRO | Annual | Festuca/Vulpia |
| *Festuca pratensis* | FPRA | Perennial | Festuca/Vulpia |

The dataset contains the 1500 most highly expressed HOGs from VST-normalized RNA-seq from the
"Live or let die?" project, stored as SummarizedExperiment objects.

## 1. Data preparation

```{r data, message = FALSE}
library(rcomplex)
library(SummarizedExperiment)
library(dplyr, warn.conflicts = FALSE)
library(tidyr)
library(purrr)
library(tibble)

se_list <- readRDS(system.file("extdata", "pooideae_vignette.rds",
                                package = "rcomplex"))
```

We organise all per-species data in a single nested tibble. Each column
accumulates one processing stage — the `tribble` defines species metadata,
then `mutate` + `map` appends the SE objects, paralog reduction, and
network construction as list-columns:

### Paralog reduction

Multi-copy gene families often contain recent duplicates with nearly
identical expression profiles. These inflate the combinatorial search
space for clique detection without adding information.
`reduce_orthogroups()` merges correlated paralogs within each HOG
using Ward.D2 hierarchical clustering on Pearson correlation distance.
Paralogs above the correlation threshold (default 0.7) are replaced
by their averaged expression; subfunctionalized copies with distinct
expression programs are preserved:

```{r species-data}
species_data <- tribble(
  ~species, ~trait,
  "BDIS",   "annual",
  "BSYL",   "perennial",
  "HVUL",   "annual",
  "HJUB",   "perennial",
  "BMAX",   "annual",
  "BMED",   "perennial",
  "VBRO",   "annual",
  "FPRA",   "perennial"
) |>
  mutate(
    se = se_list[species],
    reduced = map(se, \(s) {
      expr <- SummarizedExperiment::assay(s, 1)
      rd <- SummarizedExperiment::rowData(s)
      ortho_sp <- tibble(gene = rownames(s),
                         hog = as.character(rd$hog)) |>
        filter(!is.na(hog))
      reduce_orthogroups(expr, ortho_sp, gene_col = "gene",
                         cor_threshold = 0.7)
    })
  )

species_data |>
  select(species, reduced) |>
  pwalk(\(species, reduced)
    cat(species, ":", reduced$n_original, "->", reduced$n_reduced, "genes",
        "(", reduced$n_merged, "merged )\n"))
```

### Network construction

Build co-expression networks from the reduced expression matrices
using Mutual Rank (MR) normalization at 3% density:

```{r networks}
species_data <- species_data |>
  mutate(network = map(reduced, \(r)
    compute_network(r$expr_matrix, norm_method = "MR", density = 0.03)))

# Named lists for functions that expect them
species  <- pull(species_data, species)
trait_vec <- deframe(select(species_data, species, trait))
networks <- deframe(select(species_data, species, network))
```

### Ortholog extraction

Extract ortholog pairs from HOG annotations in rowData. After
paralog reduction, gene names may have changed (merged representatives),
so `prepare_orthologs()` maps gene names through each species'
reduction and deduplicates:

```{r orthologs}
reductions <- deframe(select(species_data, species, reduced))
orthologs <- prepare_orthologs(se_list, reductions)

cat(nrow(orthologs), "ortholog pairs across",
    length(utils::combn(species, 2, simplify = FALSE)), "species pairs\n")
```

## 2. Pairwise comparison and HOG-level testing

For each of the C(8,2) = 28 species pairs, we run two analyses:

1. **Neighborhood comparison** (`compare_neighborhoods()`): hypergeometric
   test for overlap of co-expression neighborhoods, producing per-pair
   p-values and fold-enrichment effect sizes.
2. **HOG-level permutation test** (`permutation_hog_test()`): gene-identity
   permutation within each HOG. This is necessary because pair-level tests
   within a HOG are correlated (shared neighborhoods), making Fisher's
   method or Storey q-values anti-conservative. The permutation p-value is
   exact regardless of the dependency structure.

With `use_torch = TRUE`, the fold-enrichment matrix is precomputed on
GPU, reducing each permutation to fast table lookups.

The HOG-level q-values determine which edges are co-expressologs
(conserved co-expression relationships):

```{r comparison}
n_cores <- min(4L, parallel::detectCores(logical = FALSE))

edges <- find_coexpressologs(networks, orthologs,
  method = "permutation", use_torch = TRUE, n_cores = n_cores)

cat(nrow(edges), "edges:", sum(edges$type == "conserved"), "conserved,",
    sum(edges$type == "ns"), "non-significant\n")
```

For quick exploration without GPU, the `"analytical"` method uses
pair-level Storey q-values (less statistically rigorous for multi-copy
HOGs, but much faster):

```{r analytical-edges}
edges_analytical <- find_coexpressologs(networks, orthologs,
  method = "analytical")
cat("Analytical:", sum(edges_analytical$type == "conserved"), "conserved edges\n")
cat("Permutation:", sum(edges$type == "conserved"), "conserved edges\n")
```

## 3. Module-level analysis

To identify regulatory signatures of parallel life-cycle evolution, we
detect modules in all eight species and compare across each of the four
phylogenetic pairs. Module-level conservation reveals whether the same
co-regulatory programs are used independently in annual lineages.

```{r phylo-pairs}
phylo_pairs <- tribble(
  ~pair_name,        ~annual, ~perennial,
  "Brachypodium",    "BDIS",  "BSYL",
  "Hordeum",         "HVUL",  "HJUB",
  "Briza",           "BMAX",  "BMED",
  "Festuca/Vulpia",  "VBRO",  "FPRA"
)
```

### Testing for community structure (K = 1 null)

Before partitioning, we test whether each network has significant
community structure at all. Passing a vector of resolutions triggers
multi-resolution consensus clustering (Jeub *et al.*, 2018) with a
permutation-based K = 1 test: gene labels on a degree-preserving
rewired null graph are shuffled and Leiden is re-run to build a null
distribution of the spectral norm of the excess co-classification matrix.
Early stopping terminates the permutation loop once the p-value
is decisively above or below `alpha_k1`, so networks with clear
structure typically complete in ~20 permutations rather than 100:

```{r k1-test}
resolutions <- seq(0.25, 2.5, by = 0.25)

species_data <- species_data |>
  mutate(modules = map(network, \(net)
    detect_modules(net, method = "leiden",
                   resolution = resolutions,
                   objective_function = "modularity", seed = 42,
                   test_k1 = TRUE, n_perm_k1 = 100L,
                   max_consensus_iter = 1000L)))

modules <- deframe(select(species_data, species, modules))

species_data |>
  transmute(
    species, trait,
    lambda_obs       = map_dbl(modules, \(m) m$k1_test$lambda_obs),
    p_value          = map_dbl(modules, \(m) m$k1_test$p_value),
    has_structure    = map_lgl(modules, \(m) m$k1_test$has_structure),
    n_perm_completed = map_int(modules, \(m) m$k1_test$n_perm_completed),
    n_modules        = map_int(modules, \(m) m$n_modules)
  ) |>
  show_table(caption = "K = 1 community structure test (all species)")
```

The resolution scan shows how partition granularity varies across
resolutions and how consecutive partitions agree (ARI):

```{r resolution-scan}
show_table(modules$BDIS$resolution_scan, caption = "Resolution scan (BDIS)")
```

### Module comparison across phylogenetic pairs

`compare_modules_paired()` wraps the full comparison pipeline into a
single call: for each pair it runs Jaccard permutation (shuffling the
ortholog mapping), classifies modules as conserved / species-specific,
tags group-specificity from the `group` vector, and auto-coarsens the
finer-grained partition when the module-count ratio exceeds 2-fold:

```{r module-comparison}
mod_results <- compare_modules_paired(
  modules, orthologs,
  pairs = phylo_pairs |> transmute(sp1 = annual, sp2 = perennial, pair_name),
  group = trait_vec,
  matched_scale = TRUE, coarsen_ratio = 2,
  method = "jaccard", n_cores = n_cores
)

show_table(mod_results$module_counts,
           caption = "Module count ratio per pair")
show_table(mod_results$summary,
           caption = "Module conservation: natural vs matched scale")
```

Do any HOGs consistently land in annual-specific modules across
independent lineages? That would be strong evidence for parallel
re-use of regulatory programs. `tag_permutation()` tests this
formally: trait labels are shuffled across species (preserving
marginal frequencies), and HOG recurrence is recomputed under
each permuted labelling. A pair contributes HOGs only when exactly
one of its species carries the target trait, so the null correctly
accounts for the paired design:

```{r module-parallel}
parallel_test <- tag_permutation(
  mod_results$classification, modules, orthologs,
  pairs = phylo_pairs |> transmute(sp1 = annual, sp2 = perennial, pair_name),
  group = trait_vec, target_group = "annual",
  n_perm = 1000L, min_recurrence = 2L
)

parallel_test$recurrence_table |>
  count(n_pairs, name = "n_hogs") |>
  show_table(caption = "HOG recurrence in annual-specific modules across pairs")

cat("HOGs recurring in >= 2 pairs:", parallel_test$observed,
    "(permutation p =", sprintf("%.4f", parallel_test$p_value), ")\n")
```

### Hub genes and trait conservation

Identify hub genes within each species' modules, then classify each HOG
by whether it acts as a hub across trait groups. **Weighted degree** (the
default) is the recommended primary centrality for co-expression networks
-- it directly measures total co-expression strength to module neighbours,
which is the standard definition of a hub gene. All three centrality
measures (degree, betweenness, eigenvector) are reported regardless of
which is chosen as primary.

To disentangle centrality from regulatory function,
`characterize_hubs()` adds two metrics: **bridge fraction** (the
share of a hub's co-expression edges reaching genes in other
modules -- high values indicate a regulatory coordinator rather
than an intra-module effector) and **betweenness/degree ratio**
(importance for information flow relative to raw connectivity).
When an expression matrix is provided, it also computes the
**coefficient of variation** -- regulators tend to show higher
expression variability than their targets because they drive
transcriptional changes.

```{r module-hubs}
species_data <- species_data |>
  mutate(hubs = pmap(list(species, modules, network, reduced),
    \(sp, mod, net, red)
      identify_module_hubs(mod, net, orthologs) |>
        characterize_hubs(mod, expr = red$expr_matrix)))

species_data |>
  transmute(species, trait,
            n_hubs  = map_int(hubs, \(h) sum(h$is_hub)),
            n_genes = map_int(hubs, nrow)) |>
  show_table(caption = "Hub genes per species")
```

```{r hub-conservation}
hub_class <- classify_hub_conservation(
  deframe(select(species_data, species, hubs)),
  trait_vec,
  module_comparisons = mod_results$raw
)

hub_class |>
  filter(classification != "non_hub") |>
  count(classification) |>
  show_table(caption = "Hub conservation across all species")
```

The regulatory characterization helps separate potential regulators
from highly-connected effectors. We aggregate `bridge_fraction`,
`bt_degree_ratio`, and expression `cv` to the HOG level:

```{r hub-regulatory}
hub_regulatory <- species_data |>
  select(species, trait, hubs) |>
  unnest(hubs) |>
  filter(is_hub, !is.na(hog)) |>
  summarise(
    max_bridge = max(bridge_fraction, na.rm = TRUE),
    mean_bt_deg = mean(bt_degree_ratio, na.rm = TRUE),
    mean_cv = if (all(is.na(cv))) NA_real_ else mean(cv, na.rm = TRUE),
    .by = hog
  )

hub_class |>
  filter(classification != "non_hub") |>
  left_join(hub_regulatory, by = "hog") |>
  select(hog, classification, n_species_hub, max_centrality,
         max_bridge, mean_bt_deg, mean_cv) |>
  arrange(desc(max_bridge), desc(mean_bt_deg)) |>
  show_table(caption = "Hub regulatory characterization (sorted by bridge fraction)")
```

HOGs that are hubs specifically in annual (or perennial) species -- the
`annual_specific_hub`, `perennial_specific_hub`, and `rewired_hub`
classes -- are candidates for parallel regulatory evolution. Ranking
by `bridge_fraction` and `bt_degree_ratio` prioritizes those most
likely to be true regulators:

```{r hub-trait-specific}
hub_class |>
  filter(classification %in% c("annual_specific_hub",
                                "perennial_specific_hub",
                                "rewired_hub")) |>
  left_join(hub_regulatory, by = "hog") |>
  select(hog, classification, n_species_hub, hub_trait_groups,
         max_centrality, max_bridge, mean_bt_deg, mean_cv) |>
  arrange(desc(max_bridge)) |>
  show_table(caption = "Trait-specific and rewired hubs (ranked by regulatory potential)")
```

### Co-expression partners of hub HOGs

After identifying trait-specific hubs, the natural question is: "Which
other HOGs co-express with this hub across species?"
`get_coexpressed_hogs()` resolves a HOG to its gene(s) in each species,
extracts co-expression neighbors from the network, maps them back to
HOGs, and aggregates across species. The `coexpressed_traits` column
makes trait-group filtering straightforward:

```{r hub-partners}
candidate <- hub_class |>
  filter(classification %in% c("annual_specific_hub",
                                "perennial_specific_hub",
                                "rewired_hub", "conserved_hub")) |>
  slice_max(n_species_hub, n = 1, with_ties = FALSE) |>
  pull(hog)

if (length(candidate) == 1) {
  partners <- get_coexpressed_hogs(
    candidate, networks, orthologs,
    species_trait = trait_vec, min_species = 2L,
    edges = edges
  )

  cat(candidate, ":", nrow(partners), "partner HOGs in >= 2 species —",
      sum(partners$coexpressed_traits == "annual", na.rm = TRUE), "annual-only,",
      sum(grepl(",", partners$coexpressed_traits), na.rm = TRUE), "cross-trait\n")

  partners |>
    slice_head(n = 20) |>
    show_table(caption = paste("Top co-expression partners of", candidate))
} else {
  cat("No trait-specific or conserved hubs found\n")
}
```

## 4. Edge robustness: density sweep

**Network density** controls how many edges are retained in the
co-expression network. At 3% density, only the top 3% of all
pairwise MR values pass the threshold -- these are the strongest
co-expression relationships.

We sweep over 15 density levels from 0.8x to 1.5x the baseline
threshold, testing both directions:

- **Tighter** (0.8x--0.95x, effective density ~2.4%--2.85%): does
  the signal survive a 5--20% reduction in edges? Surviving
  co-expressologs are strong, not marginal.
- **Looser** (1.05x--1.5x, effective density ~3.15%--4.5%): is the
  signal specific to this density range, or does it persist as new
  edges are added? Cliques that only appear at very loose thresholds
  are likely noise from extra edges.

The 3% density choice is somewhat arbitrary. Sweeping a wide range
around it reveals sensitivity to that choice -- a truly robust
co-expressolog should be present across a range of densities, not
just at one particular cutoff.

The permutation pipeline is re-run at each level because network
neighborhoods change with the threshold, altering the HOG-level
q-values:

```{r density-sweep}
sweep <- density_sweep(networks, orthologs,
  multipliers = seq(0.8, 1.5, by = 0.05),
  method = "permutation", use_torch = TRUE, n_cores = n_cores)

# Edge survival across density levels
sweep |>
  transmute(
    Multiplier = paste0(multiplier, "x"),
    Eff_density = sprintf("%.2f%%", eff_density * 100),
    Conserved_edges = n_significant,
    Baseline = if_else(multiplier == 1, "***", "")
  ) |>
  show_table(caption = "Conserved edges across density levels")
```

The mean effective density masks species-level variation. Because
each species has a different MR distribution shape, a uniform
threshold multiplier produces different effective densities across
species:

```{r species-density-variation}
baseline_densities <- sweep |>
  filter(multiplier == 1) |>
  pull(species_densities) |>
  pluck(1)

tibble(
  species = names(baseline_densities),
  trait = trait_vec[names(baseline_densities)],
  eff_density = sprintf("%.2f%%", baseline_densities * 100)
) |>
  show_table(caption = "Per-species effective density at baseline (1.0x)")
```

The edge count may not decrease monotonically because the permutation
test operates at the HOG level: tighter neighborhoods can sharpen the
signal for some HOGs while eliminating it for others. Local fluctuations
reflect the discrete nature of the permutation q-values. The
interpretation is asymmetric: surviving tighter thresholds indicates
strong signal, while surviving looser thresholds indicates specificity
(not merely noise from additional edges).

**Analytical alternative: `clique_persistence()`.**
The density sweep above is rigorous but expensive — it re-runs the
full permutation pipeline at each threshold level. For a quick estimate,
`clique_persistence()` computes the ratio of each co-expressolog edge's
MR value to the species' density threshold. A persistence of 1.5 means
the weakest supporting edge is 50% above threshold and would survive a
1.5x tightening. This is a principled analytical proxy: it correctly
identifies the weakest link in each clique's co-expressolog support.
However, it cannot detect that tighter neighborhoods may change which
HOGs reach significance in the permutation test, or that new cliques
may form from HOGs that become significant at stricter thresholds.
Use persistence for screening or when GPU is unavailable; use the full
sweep for publication-quality robustness assessment.

## 5. Clique detection and structural stability

**Phylogenetic structure of pairwise comparisons.** The
co-expressolog edge set contains C(8,2) = 28 species pairs, mixing
three types: (1) **focal pairs** from the same genus (4 annual vs
perennial sister pairs), (2) **within-trait pairs** comparing two
annuals or two perennials from different genera (12 pairs), and (3)
**cross-trait/cross-genus pairs** (12 pairs). Cliques built on all
28 pairs can span phylogenetically non-independent comparisons.
`classify_cliques()` handles this by separating within-trait from
cross-trait cliques in the waterfall, but cliques are not weighted
by phylogenetic distance. For the current dataset this is acceptable:
each genus contributes exactly one species per trait, limiting
phylogenetic confounding.

Using edges from the baseline density level, we find cliques and
test their structural stability via leave-k-out jackknife. The
jackknife removes k species at a time and checks whether each
clique's gene assignments are preserved (purely structural,
trait-agnostic). Trait annotations are added post-hoc:

```{r cliques-stability}
# Use the baseline (1.0x) edges for clique detection and stability
robust_edges <- sweep |> filter(multiplier == 1) |> pull(edges) |> pluck(1)
cat("Using", sum(robust_edges$type == "conserved"), "conserved edges at",
    sprintf("%.2f%%",
            (sweep |> filter(multiplier == 1) |> pull(eff_density)) * 100),
    "density\n\n")

# Find cliques
cliques <- find_cliques(robust_edges, species, min_species = 2L)
cat(nrow(cliques), "cliques\n")

cliques |>
  count(n_species) |>
  show_table(pageLength = 10, caption = "Cliques by number of species")
```

Cliques with `n_species = 2` are single conserved edges between one
pair of species -- the weakest form of co-expression conservation.
The most compelling evidence comes from cliques spanning 3 or more
species, as these require coordinated conservation across multiple
independent lineages.

The conserved edges feeding into clique detection come from three
types of species pair:

```{r pair-type-breakdown}
focal_set <- paste(phylo_pairs$annual, phylo_pairs$perennial)
robust_edges |>
  filter(type == "conserved") |>
  mutate(
    t1 = trait_vec[species1], t2 = trait_vec[species2],
    pair_type = case_when(
      paste(species1, species2) %in% focal_set |
        paste(species2, species1) %in% focal_set ~ "focal (sister pair)",
      t1 == t2 ~ "within-trait",
      TRUE ~ "cross-trait/cross-genus"
    )
  ) |>
  count(pair_type) |>
  show_table(caption = "Conserved edges by pair type")
```

```{r stability}
# Jackknife stability (all k from 1 to N-2)
max_k <- length(species) - 2L
stab <- clique_stability(robust_edges, species,
                          species_trait = trait_vec,
                          all_species = species,
                          full_cliques = cliques,
                          max_k = max_k)
cat("Cliques tested:", length(stab$stability_class), "\n")
cat("Structurally stable at k=1:", sum(stab$stability_class >= 1L), "\n")

tibble(stability_class = stab$stability_class) |>
  mutate(stability_class = factor(stability_class, levels = 0:max_k)) |>
  count(stability_class, .drop = FALSE) |>
  show_table(pageLength = 10, caption = "Stability class distribution")
```

## 6. HOG classification with robustness ranking

Classify each HOG by its conservation pattern, then rank within each
category by robustness. The waterfall assigns each HOG to exactly one
class: complete > partial > differentiated > trait_specific > unclassified.

```{r classify}
result <- classify_cliques(robust_edges, species, trait_vec,
                            min_species = 2L,
                            stability = stab)

result |>
  count(classification) |>
  show_table(pageLength = 5, caption = "HOG classification summary")
```

Restricting to cliques spanning 3 or more species (excluding single-edge
2-species signals) gives a clearer picture of multi-lineage conservation:

```{r focal-classification}
result |>
  filter(n_species >= 3 | is.na(n_species)) |>
  count(classification) |>
  show_table(pageLength = 5,
             caption = "HOG classification (cliques >= 3 species only)")
```

- **complete**: all 8 species form a clique -- deeply conserved across
  all Pooideae
- **partial**: clique spans both annual and perennial species, but not
  all 8
- **differentiated**: annual species have a clique AND perennial species
  have a clique, but no cross-group conservation -- lineage-specific
  co-expression programs
- **trait_specific**: only annuals or only perennials have a clique
- **unclassified**: no conservation signal detected

The robustness ranking comes from combining density sweep survival
with stability class. HOGs classified at a stricter density level
with high stability class are the strongest findings:

```{r ranking}
# Count how many density levels each HOG's edges survive
hog_survival <- sweep |>
  transmute(multiplier,
            hog = map(edges, \(e)
              e |> filter(type == "conserved") |> distinct(hog) |> pull(hog))) |>
  unnest(hog) |>
  count(hog, name = "n_densities_survived")

result <- result |>
  left_join(hog_survival, by = "hog") |>
  mutate(n_densities_survived = replace_na(n_densities_survived, 0L))

# Annotate trait composition from the best clique per HOG
hog_traits <- cliques |>
  group_by(hog) |>
  slice_min(mean_q, n = 1, with_ties = FALSE) |>
  ungroup() |>
  pivot_longer(all_of(species), names_to = "sp", values_to = "gene") |>
  filter(!is.na(gene)) |>
  summarise(
    n_annual = sum(trait_vec[sp] == "annual"),
    n_perennial = sum(trait_vec[sp] == "perennial"),
    .by = hog
  )

result <- left_join(result, hog_traits, by = "hog")

# Show ranked HOGs per category
show_cols <- c("hog", "n_species", "n_annual", "n_perennial",
               "best_mean_q", "stability_class", "n_densities_survived")
show_cols <- intersect(show_cols, names(result))
```

Every clique now has a `stability_class` -- the highest k at which it
survives ALL species-removal subsets. This is a purely structural measure.
Trait annotations (from the `traits` column in the stability output) are
orthogonal: they describe which life-cycle groups each clique spans, not
whether the clique is robust.

### {.tabset}

#### Partial

```{r partial-table}
show_classification(result, "partial", show_cols,
                    "Partial HOGs (cross-trait cliques)")
```

#### Trait-specific

```{r trait-specific-table}
show_classification(result, "trait_specific", show_cols,
                    "Trait-specific HOGs (one lineage only)")
```

#### Differentiated

```{r differentiated-table}
show_classification(result, "differentiated", show_cols,
                    "Differentiated HOGs (within-group conservation, cross-group divergence)")
```

#### Complete

```{r complete-table}
show_classification(result, "complete", show_cols,
                    "Complete HOGs (all 8 species)")
```

###

### Clique robustness diagnostics

The density sweep and jackknife stability above assess whether cliques
survive threshold changes and species removal. Several additional
diagnostics quantify clique quality at a finer grain.

**Edge-level quality metrics.** `find_cliques()` now returns three
columns alongside each clique: `intensity` (Onnela intensity -- the
geometric mean of `1 - q.value` across present edges), `coherence`
(intensity divided by the arithmetic mean, equal to 1 when all edge
weights are equal), and `min_effect_size` (the bottleneck enrichment
across present edges). High intensity with high coherence indicates
uniformly strong conservation; high intensity with low coherence
flags cliques carried by one or two dominant edges. The
`cost_weights` parameter controls gene-assignment ranking: the
default `c(q = 1, effect = 0)` minimises mean q-value, while
`c(q = 1, effect = 1)` uses a composite cost
`q * mean_q - effect * mean_effect` that also rewards large
effect sizes:

```{r intensity-coherence, eval = FALSE}
# Intensity and coherence are already in cliques from find_cliques()
cliques |>
  select(hog, n_species, mean_q, intensity, coherence, min_effect_size) |>
  filter(n_species >= 3) |>
  arrange(desc(intensity)) |>
  head(20) |>
  show_table(caption = "Top cliques by Onnela intensity (>= 3 species)")

# Composite cost: rank gene assignments by both q-value and effect size
cliques_composite <- find_cliques(robust_edges, species, min_species = 2L,
                                   cost_weights = c(q = 1, effect = 0.5))
```

**Jaccard overlap in edge output.** `comparison_to_edges()` (and
therefore `find_coexpressologs()`) now includes a `jaccard` column:
the geometric mean of the two directional Jaccard indices of
neighborhood overlap. This complements the hypergeometric q-value
with a non-parametric measure of neighborhood similarity (range
0--1). Large effect size with low Jaccard suggests enrichment driven
by a small shared subset rather than broad neighborhood overlap:

```{r edge-jaccard, eval = FALSE}
robust_edges |>
  filter(type == "conserved") |>
  select(hog, gene1, gene2, species1, species2, q.value, effect_size, jaccard) |>
  arrange(desc(jaccard)) |>
  head(20) |>
  show_table(caption = "Top conserved edges by Jaccard overlap")
```

**Birth/death persistence.** `clique_threshold_sweep()` returns a
`$persistence` element with one row per baseline clique. Each row
records the `birth` (lowest multiplier where the clique exists),
`death` (first multiplier above birth where it disappears), and
`persistence` (death minus birth). A persistence of `NA` means the
clique survived all tested multipliers. Unlike the density sweep
survival count, persistence captures the continuous range of
thresholds over which each clique lives:

```{r birth-death, eval = FALSE}
sweep_result <- clique_threshold_sweep(
  cliques, species, networks, orthologs,
  multipliers = c(1.5, 2, 3, 5, 10),
  min_species = 2L, n_cores = n_cores
)

sweep_result$persistence |>
  arrange(desc(persistence)) |>
  head(20) |>
  show_table(caption = "Clique birth/death persistence")
```

**Perturbation test.** `clique_perturbation_test()` adds Gaussian
noise to MR scores in each species network, re-thresholds, re-runs
the comparison-to-cliques pipeline, and measures how often each
baseline clique survives (Jaccard >= threshold to its perturbed
counterpart). This tests robustness to measurement noise without
recomputing correlations from scratch. Cliques with high
`survival_rate` are insensitive to small perturbations in edge
weights:

```{r perturbation, eval = FALSE}
pert <- clique_perturbation_test(
  cliques, species, networks, orthologs,
  n_boot = 100, noise_sd = 0.1,
  min_species = 2L, jaccard_threshold = 0.5,
  n_cores = n_cores, seed = 42
)

pert |>
  arrange(desc(survival_rate)) |>
  head(20) |>
  show_table(caption = "Clique perturbation survival (100 bootstrap iterations)")
```

**Intensity Z-score.** `clique_intensity_test()` builds a null
distribution by permuting the ortholog mapping (globally shuffling
Species2 genes across all rows, destroying both the specific mapping
and within-HOG gene grouping). For each permutation, cliques are
re-detected and the intensity of the best-matching clique (by
Jaccard) is recorded. The Z-score and empirical p-value test whether
the observed gene-to-gene correspondence produces stronger
co-expression conservation than random mappings:

```{r intensity-test, eval = FALSE}
intens <- clique_intensity_test(
  cliques, species, networks, orthologs,
  n_perm = 500, min_species = 2L,
  n_cores = n_cores, seed = 42, edges = robust_edges
)

intens |>
  filter(p_value < 0.05) |>
  arrange(desc(z_score)) |>
  head(20) |>
  show_table(caption = "Cliques with significantly elevated intensity (p < 0.05)")
```

Together, these diagnostics address complementary robustness
questions: density sweep tests threshold sensitivity, jackknife
tests species-removal stability, perturbation tests measurement
noise, and the intensity Z-score tests whether the ortholog mapping
itself drives the signal. Cliques passing all four are the strongest
candidates for biologically meaningful co-expression conservation.

## 7. Biological interpretation

The central question is whether independent annual Pooideae lineages
draw on a shared, latent regulatory architecture inherited from
perennial ancestors. The analyses above provide several lines of
evidence:

```{r synthesis}
n_complete <- sum(result$classification == "complete", na.rm = TRUE)
n_partial <- sum(result$classification == "partial", na.rm = TRUE)
n_differentiated <- sum(result$classification == "differentiated", na.rm = TRUE)
n_trait_specific <- sum(result$classification == "trait_specific", na.rm = TRUE)

# Robust candidates: survive >= half the density levels and stability >= 1
n_robust <- sum(
  result$n_densities_survived >= ceiling(nrow(sweep) / 2) &
    !is.na(result$stability_class) & result$stability_class >= 1L,
  na.rm = TRUE
)

cat("Complete (all 8 species):", n_complete, "\n")
cat("Partial (cross-trait, < 8):", n_partial, "\n")
cat("Differentiated (independent annual + perennial cliques):",
    n_differentiated, "\n")
cat("Trait-specific (one lineage only):", n_trait_specific, "\n")
cat("Robust candidates (>= 50% density survival + stability >= 1):",
    n_robust, "\n\n")
cat("Parallel module signal: ", parallel_test$observed,
    " HOGs recur in annual-specific modules across >= 2 pairs (p = ",
    sprintf("%.4f", parallel_test$p_value), ")\n", sep = "")
```

**Differentiated HOGs** are the strongest candidates for regulatory
divergence: the same ortholog group has co-expression relationships in
both annuals and perennials, but these programs are incompatible (no
cross-group clique). This pattern is consistent with a shared ancestral
regulatory architecture that has been independently remodelled in each
annual lineage.

**Trait-specific HOGs** (annual-only or perennial-only cliques) may
encode lineage-specific regulatory modules. The formal permutation
test above addresses whether annual-specific modules share HOGs
across independent lineages more than expected by chance. A significant
result supports the hypothesis that annuality draws on a shared
regulatory toolkit rather than arising from independent *de novo*
rewiring in each lineage.

**Robustness ranking** combines density sweep survival with jackknife
stability class. HOGs classified at tighter density levels with high
stability represent the strongest findings for experimental follow-up.

**Regulatory characterization** helps prioritize among hub
candidates. Hubs with high `bridge_fraction` connect across modules,
a hallmark of regulatory coordinators (TFs, signalling genes). High
`bt_degree_ratio` indicates importance for information flow relative
to raw connectivity. High expression `cv` suggests the gene drives
transcriptional changes rather than passively responding. Candidates
scoring high on all three metrics are the strongest regulatory hub
candidates, especially when they also carry known DNA-binding or
signalling domains (testable via an annotation overlay with
PlantTFDB or InterPro).

**Caveats.** Co-expression conservation is necessary but not sufficient
evidence for regulatory conservation. Shared co-expression could
reflect shared ancestry (phylogenetic inertia) rather than convergent
regulatory function. The paired phylogenetic design partially controls
for this -- each pair is an independent comparison within a genus --
but formal ancestral state reconstruction of co-expression networks
would be needed to fully distinguish these scenarios. The candidate
HOGs identified here should be validated with functional annotation
(GO enrichment, transcription factor binding motifs) and, where
available, tissue- or stage-specific expression data.

## Session info

```{r sessioninfo}
sessionInfo()
```
