---
title: "Comparative co-expression analysis with rcomplex"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparative co-expression analysis with rcomplex}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_se <- requireNamespace("SummarizedExperiment", quietly = TRUE)
has_torch <- requireNamespace("torch", quietly = TRUE) &&
  torch::torch_is_installed() &&
  tryCatch(
    {
      torch::torch_randn(1)
      TRUE
    },
    error = function(e) FALSE
  )
has_tidy <- requireNamespace("purrr", quietly = TRUE) &&
  requireNamespace("tibble", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  eval = has_se && has_torch && has_tidy
)

# Helper for interactive tables (sortable by one or multiple columns)
show_table <- function(df, pageLength = 10, ...) {  # nolint
  if (requireNamespace("DT", quietly = TRUE)) {
    DT::datatable(df,
      rownames = FALSE,
      options = list(
        pageLength = pageLength, scrollX = TRUE,
        ordering = TRUE, orderMulti = TRUE
      ),
      ...
    )
  } else {
    knitr::kable(df)
  }
}

# Helper for ranked classification tables
show_classification <- function(df, class, cols, caption) {
  sub <- df |>
    dplyr::filter(.data$classification == class) |>
    dplyr::select(dplyr::any_of(cols)) |>
    dplyr::arrange(
      dplyr::desc(.data$n_species), dplyr::desc(.data$n_densities_survived),
      .data$best_mean_q
    )
  show_table(sub, pageLength = 10, caption = caption)
}
```

This vignette demonstrates a complete rcomplex analysis using expression
data from eight Pooideae grass species spanning four phylogenetic pairs,
each containing one annual and one perennial species:

| Species | Code | Life cycle | Pair |
|---------|------|------------|------|
| *Brachypodium distachyon* | BDIS | Annual | Brachypodium |
| *Brachypodium sylvaticum* | BSYL | Perennial | Brachypodium |
| *Hordeum vulgare* | HVUL | Annual | Hordeum |
| *Hordeum jubatum* | HJUB | Perennial | Hordeum |
| *Briza maxima* | BMAX | Annual | Briza |
| *Briza media* (crossmapped to BMAX) | BMED | Perennial | Briza |
| *Vulpia bromoides* | VBRO | Annual | Festuca/Vulpia |
| *Festuca pratensis* | FPRA | Perennial | Festuca/Vulpia |

The dataset contains the 1500 most highly expressed HOGs from VST-normalized RNA-seq from the
"Live or let die?" project, stored as SummarizedExperiment objects.

## 1. Data preparation

```{r data, message = FALSE}
library(rcomplex)
library(SummarizedExperiment)
library(dplyr, warn.conflicts = FALSE)
library(tidyr)
library(purrr)
library(tibble)

se_list <- readRDS(system.file("extdata", "pooideae_vignette.rds",
  package = "rcomplex"
))
```

We organise all per-species data in a single nested tibble. Each column
accumulates one processing stage — the `tribble` defines species metadata,
then `mutate` + `map` appends the SE objects, paralog reduction, and
network construction as list-columns:

### Paralog reduction

Multi-copy gene families often contain recent duplicates with nearly
identical expression profiles. These inflate the combinatorial search
space for clique detection without adding information.
`reduce_orthogroups()` merges correlated paralogs within each HOG
using Ward.D2 hierarchical clustering on Pearson correlation distance.
Paralogs above the correlation threshold (default 0.7) are replaced
by their averaged expression; subfunctionalized copies with distinct
expression programs are preserved:

```{r species-data}
species_data <- tribble(
  ~species, ~trait,
  "BDIS",   "annual",
  "BSYL",   "perennial",
  "HVUL",   "annual",
  "HJUB",   "perennial",
  "BMAX",   "annual",
  "BMED",   "perennial",
  "VBRO",   "annual",
  "FPRA",   "perennial"
) |>
  mutate(
    se = se_list[species],
    reduced = map(se, \(s) {
      expr <- SummarizedExperiment::assay(s, 1)
      rd <- SummarizedExperiment::rowData(s)
      ortho_sp <- tibble(
        gene = rownames(s),
        hog = as.character(rd$hog)
      ) |>
        filter(!is.na(hog))
      reduce_orthogroups(expr, ortho_sp,
        gene_col = "gene",
        cor_threshold = 0.7
      )
    })
  )

species_data |>
  select(species, reduced) |>
  pwalk(\(species, reduced) {
    cat(
      species, ":", reduced$n_original, "->", reduced$n_reduced, "genes",
      "(", reduced$n_merged, "merged )\n"
    )
  })
```

### Network construction

Build co-expression networks from the reduced expression matrices
using Mutual Rank (MR) normalization at 3% density. The sparse store
is built at 6% (`store_density = 0.06`) to leave headroom for the
density sweep in section 4, which loosens the networks below the
default 5% store:

```{r networks}
species_data <- species_data |>
  mutate(network = map(reduced, \(r) {
    compute_network(r$expr_matrix,
      norm_method = "MR", density = 0.03,
      store_density = 0.06
    )
  }))

# Named lists for functions that expect them
species <- pull(species_data, species)
trait_vec <- deframe(select(species_data, species, trait))
networks <- deframe(select(species_data, species, network))
```

### Ortholog extraction

Extract ortholog pairs from HOG annotations in rowData. After
paralog reduction, gene names may have changed (merged representatives),
so `prepare_orthologs()` maps gene names through each species'
reduction and deduplicates:

```{r orthologs}
reductions <- deframe(select(species_data, species, reduced))
orthologs <- prepare_orthologs(se_list, reductions)

cat(
  nrow(orthologs), "ortholog pairs across",
  length(utils::combn(species, 2, simplify = FALSE)), "species pairs\n"
)
```

## 2. Pairwise comparison and HOG-level testing

For each of the C(8,2) = 28 species pairs, we run two analyses:

1. **Neighborhood comparison** (`compare_neighborhoods()`): hypergeometric
   test for overlap of co-expression neighborhoods, producing per-pair
   p-values and fold-enrichment effect sizes.
2. **HOG-level permutation test** (`permutation_hog_test()`): gene-identity
   permutation within each HOG. This is necessary because pair-level tests
   within a HOG are correlated (shared neighborhoods), making Fisher's
   method or Storey q-values anti-conservative. The permutation p-value is
   exact regardless of the dependency structure.

With `use_torch = TRUE`, the fold-enrichment matrix is precomputed on
GPU, reducing each permutation to fast table lookups.

The HOG-level q-values determine which edges are co-expressologs
(conserved co-expression relationships):

```{r comparison}
set.seed(1)
n_cores <- min(4L, parallel::detectCores(logical = FALSE))

edges <- find_coexpressologs(networks, orthologs,
  method = "permutation", use_torch = TRUE, n_cores = n_cores
)

cat(
  nrow(edges), "edges:", sum(edges$type == "conserved"), "conserved,",
  sum(edges$type == "ns"), "non-significant\n"
)
```

The permutation engine draws its gene-identity permutations from the
global RNG, so the seed above pins the HOG q-values -- but only at
`n_cores = 1`. Above one core the R seed is used to seed one RNG per
worker thread, and OpenMP's `schedule(guided)` assigns HOGs to threads
dynamically, based on runtime timing. So which HOG draws from which
thread's stream is not fixed by the seed, the machine, or the core
count: two knits on the very same machine with the same `n_cores` can
still give different conserved-edge counts. Set `n_cores = 1L` if
the permutation numbers must be bit-reproducible.

For quick exploration without GPU, the `"analytical"` method uses
pair-level Storey q-values (less statistically rigorous for multi-copy
HOGs, but much faster). Since v0.2.0 its default `pi0_method =
"randomized"` estimates the null proportion pi0 from randomized
p-values. Those draws move the q-values, and with them every count
thresholded on a q-value downstream, so pass `seed` and the call becomes
reproducible on its own terms. Without a seed the draws come from the
global RNG and the result depends on whatever the session happened to
draw before.

The seed is not cosmetic here. `edges_analytical` is also the input to
the gene-graph taxonomy in section 6, and the randomized-p draw shifts
every q-value by a small amount, which moves a handful of edges across
`alpha_graph` and changes the tier counts: on this dataset
`complete_conserved` holds at 391 across seeds, but
`partial_significant` and `unclassified` move by a few cliques. Those
two numbers are only reproducible from one knit to the next because of
the `seed` below.

```{r analytical-edges}
edges_analytical <- find_coexpressologs(networks, orthologs,
  method = "analytical", seed = 1
)
cat(
  "Analytical:", sum(edges_analytical$type == "conserved"), "conserved edges\n"
)
cat("Permutation:", sum(edges$type == "conserved"), "conserved edges\n")
```

## 3. Module-level analysis

To identify regulatory signatures of parallel life-cycle evolution, we
detect modules in all eight species and test preservation across each of
the four phylogenetic pairs. Module-level conservation reveals whether the
same co-regulatory programs are used independently in annual lineages.
Preservation is directional -- whether one species' modules keep their
wiring in the partner network is a different question from the reverse --
so both directions are run for every pair, and every pair contributes two
sets of rows instead of one. Preservation is run over *all* 28 species
pairs, not only the four within-genus contrasts: the primary trait test
weighs same-trait pairs against different-trait ones, and a within-genus
table holds no same-trait pair at all. The four within-genus contrasts
are still used, by the secondary HOG-recurrence test at the end of the
section.

```{r phylo-pairs}
phylo_pairs <- tribble(
  ~pair_name,        ~annual, ~perennial,
  "Brachypodium",    "BDIS",  "BSYL",
  "Hordeum",         "HVUL",  "HJUB",
  "Briza",           "BMAX",  "BMED",
  "Festuca/Vulpia",  "VBRO",  "FPRA"
)
```

### Testing for community structure (K = 1 null)

Before partitioning, we test whether each network has significant
community structure at all. Passing a vector of resolutions triggers
multi-resolution consensus clustering (Jeub *et al.*, 2018) with a
permutation-based K = 1 test: gene labels on a degree-preserving
rewired null graph are shuffled and Leiden is re-run to build a null
distribution of the spectral norm of the excess co-classification matrix.
Early stopping terminates the permutation loop once the p-value
is decisively above or below `alpha_k1`, so networks with clear
structure typically complete in ~20 permutations rather than 100:

```{r k1-test}
resolutions <- seq(0.25, 2.5, by = 0.25)

species_data <- species_data |>
  mutate(modules = map(network, \(net) {
    detect_modules(net,
      method = "leiden",
      resolution = resolutions,
      objective_function = "modularity", seed = 42,
      test_k1 = TRUE, n_perm_k1 = 100L,
      max_consensus_iter = 1000L
    )
  }))

modules <- deframe(select(species_data, species, modules))

species_data |>
  transmute(
    species, trait,
    lambda_obs = map_dbl(modules, \(m) m$k1_test$lambda_obs),
    p_value = map_dbl(modules, \(m) m$k1_test$p_value),
    has_structure = map_lgl(modules, \(m) m$k1_test$has_structure),
    n_perm_completed = map_int(modules, \(m) m$k1_test$n_perm_completed),
    n_modules = map_int(modules, \(m) m$n_modules)
  ) |>
  show_table(caption = "K = 1 community structure test (all species)")
```

The resolution scan shows how partition granularity varies across
resolutions and how consecutive partitions agree (ARI):

```{r resolution-scan}
show_table(modules$BDIS$resolution_scan, caption = "Resolution scan (BDIS)")
```

### Module preservation across phylogenetic pairs

`preservation_paired()` wraps the full preservation pipeline into a
single call. For each pair, in both directions, it asks whether the
reference species' modules keep their wiring in the partner network:
module density (`avg.weight`) and hub identity (`cor.degree`) are each
tested against a gene-identity permutation null and combined with `pmax`,
so a module counts as preserved only when both statistics are
significant. Every module is then labelled `conserved` (significant with
`Zsummary` at or above 10), `moderate` (significant but weaker),
`diverged` (not significant), or `untested` -- a statistic was undefined,
so nothing was measured, which is not the same as a positive claim of
divergence. With a `group` vector, diverged modules are attributed to the
trait group of the species that owns them and preserved ones to
`"conserved"`, while untested modules are carried as their own level
rather than folded into a trait. There is no matched-scale table any
more: preservation projects one species' modules onto the other's network
instead of comparing two partitions, so a module-count ratio has nothing
left to correct for.

Run it over *every* pair of species, not only the four within-genus
contrasts. The extra 24 pairs are what the trait test in the next section
consumes, and it consumes them because it has nothing to work with
otherwise. Its statistic is the mean preservation of trait-*concordant*
pairs minus that of trait-*discordant* ones, and in this design every
within-genus pair is one annual against one perennial: a designated
four-contrast table is discordant end to end, so the concordant mean is
taken over an empty set and `preservation_matrix_test()` stops rather
than returning a number. Only the between-genus pairs supply
annual-annual and perennial-perennial contrasts. The cost is linear in
pairs: seven times the preservation runs.

```{r module-preservation}
# `cliques` is not built until section 5, so paralogs are resolved from
# the coexpressolog edges alone here.
mod_results <- preservation_paired(
  modules, networks, orthologs,
  pairs = all_species_pairs(names(modules)),
  group = trait_vec,
  edges = edges, n_cores = n_cores, seed = 42
)

tibble(
  species = names(modules),
  trait = trait_vec[names(modules)],
  n_modules = map_int(modules, \(m) m$n_modules)
) |>
  show_table(caption = "Modules per species")

show_table(mod_results$summary,
  caption = "Module preservation by contrast, direction and group"
)
```

### Is divergence associated with the trait?

The primary trait test asks one question of the whole preservation
matrix: **are trait-discordant species pairs the less-preserved ones?**
`preservation_matrix_test()` averages `Zsummary_std` over the
module-directions of concordant pairs (both annual, or both perennial)
and of discordant pairs (one of each), takes the difference, and refers
it to a null that relabels the species' traits. A positive statistic
means discordant pairs are less preserved, which is the biological
hypothesis, so the test is one-sided upward.

Two things make this the primary analysis rather than an extra one.

First, it uses **every** pairwise comparison instead of a designated few,
which is what gives the statistic a concordant side to average at all.
All 28 pairs are run and 24 contribute rows (the four within-genus ones
are dropped, for the reason given below); of those 24, 12 are
trait-concordant and 12 discordant. Restricted to the four within-genus
contrasts there would be no concordant pair anywhere in the table, and
the function refuses such a table outright -- with
`exclude_within_block = TRUE` because the exclusion empties it, with
`FALSE` because one side of the difference is then empty.

The two label spaces are a property of the eight species and their trait
labels, not of the pairs table, so neither of them is something the
all-pairs run buys. Splitting four annual labels among eight species
gives `choose(8, 4) = 70` free labellings, and permuting only within the
four genera gives `2^4 = 16`. What the floor is *not* is one over those
counts: renaming the two trait levels reproduces the statistic exactly,
so at least two labellings tie at the maximum and the smallest attainable
p-value is `2 / n_labellings` -- 2/70 = 0.029 free, 2/16 = 0.125 blocked.
`$free$p_attainable` and `$blocked$p_attainable` report whichever floor
actually binds, and the function warns when it exceeds 0.05.

Both nulls are reported: `p_free` over all 70 labellings, and
`p_blocked` over the 16 that permute only within a genus, which holds the
phylogeny fixed. Their agreement is the diagnostic. A free p-value far
below the blocked one would mean the larger label space bought its
resolution by breaking phylogenetic control rather than by measuring the
trait. Read `p_blocked` as that check on direction and rank only: its
floor of 0.125 puts significance out of reach for any signal, so a
non-significant blocked p-value is not evidence against the free one.

Second, it ranks on `Zsummary_std`, the standardised effect size, and
never on the q-value -- see the saturation note below for why that is not
a stylistic choice.

Within-genus pairs are dropped by default. In this design every
within-genus pair is trait-discordant and every trait-concordant pair is
between-genus, so trait status and phylogenetic distance are perfectly
confounded there, and the confound runs *against* the hypothesis: close
relatives are better preserved, which pulls those discordant rows upward
and masks a real effect. The exclusion is by genus membership, which no
relabelling changes, so the null stays valid.

```{r preservation-matrix}
genus_vec <- setNames(
  rep(phylo_pairs$pair_name, 2L),
  c(phylo_pairs$annual, phylo_pairs$perennial)
)

pmt <- preservation_matrix_test(
  mod_results$classification, trait_vec,
  block = genus_vec
)

show_table(pmt$class_means,
  caption = "Mean Zsummary_std by pair class"
)

cat(
  "mean Zsummary_std, concordant - discordant:",
  sprintf("%.3f", pmt$observed),
  "\n  p, free labelling null   :", sprintf("%.4f", pmt$p_free),
  " (floor", sprintf("%.4f", pmt$free$p_attainable), ")",
  "\n  p, within-genus null     :", sprintf("%.4f", pmt$p_blocked),
  " (floor", sprintf("%.4f", pmt$blocked$p_attainable), ")",
  # n_pairs counts the pairs left after the exclusion, while n_excluded
  # counts dropped ROWS, so the two must not be read as one sum.
  "\n  module-directions used:", pmt$n_rows, "over", pmt$n_pairs,
  "between-genus pairs",
  "\n  module-directions dropped as within-genus:", pmt$n_excluded, "\n"
)
```

Every module-direction counts once, so a heavily partitioned species
contributes more rows than a coarsely partitioned one. That is
deliberate --- the module is the unit and more modules is more evidence
--- but check `$rows_per_pair` before believing a result that one species
could be driving.

```{r preservation-matrix-rows}
show_table(pmt$rows_per_pair,
  caption = "Module-directions contributed by each species pair"
)
```

### A note on p-value saturation

A permutation p-value cannot fall below `1 / (n_perm + 1)`, because the
observed labelling is itself one of the draws. Every test whose true
p-value lies below that floor comes back holding exactly it, and
Benjamini-Hochberg maps a tied block of inputs onto a tied block of
outputs. The `preservation_paired()` call above left `n_perm` at the
`module_preservation()` default of 10000, so no p-value behind it can
fall below 1/10001 and every module whose true p-value is smaller comes
back holding exactly that. The chunk below counts the q-value block those
ties produce, a little above the p-floor once Benjamini-Hochberg has
scaled it, and a second block at exactly 1. Both counts move with
`n_perm` -- on the full eight-species set the floor ties fall from 35 at
`n_perm = 2000` to 7 at `n_perm = 20000` -- so read the numbers the chunk
prints rather than any fixed pair. What does not move is that the tied
modules are not equivalent: `Zsummary_std` spreads them over an order of
magnitude while the q-value orders none of them.

That is why every statistic above ranks on `Zsummary_std` and why `p` and
`q` appear only in the significance call. `pvalue_resolution()` puts a
number on it, so the tie count can be reported next to the p-value and a
reader can see the resolution the test actually had:

```{r preservation-saturation}
pvalue_resolution(mod_results$classification$q.value)
```

### Secondary: do the same orthogroups recur in diverged modules?

The matrix test above asks whether divergence *tracks the trait*. It says
nothing about **which** genes are involved, or whether the same ones are
involved twice. `tag_permutation()` asks that narrower question: do
particular HOGs consistently land in modules diverged in the annual
across independent lineages? A yes would be evidence for parallel re-use
of the same regulatory programs, which is a stronger and more specific
claim than the matrix test makes. It is a secondary analysis here, not a
deprecated one --- nothing else in the package answers it.

Its design is a randomised block --- the phylogenetic pair is the block,
its two species are the two levels of the trait, and the pairs are the
replication --- so the null swaps the two trait labels *within* each pair
rather than shuffling them across species. Every draw therefore keeps one
annual and one perennial per pair, which is the design that was actually
run. Contrasts that share a species are coupled, so the unit of
independence is the connected component of the graph whose nodes are
species and whose edges are contrasts. Here the four genera are disjoint,
so each is its own component, giving at most `2^4 = 16` labellings. It is
*at most* because a component whose labellings all select the same HOG
sets adds no resolution and is collapsed --- read `$n_labellings` and
`$n_swappable` rather than assuming `2^k`.

**That is also its structural limit, and it is worth stating plainly.**
The observed labelling is always one of the points in the null, so the
smallest attainable p-value is one over the number of labellings. With
four independent contrast groups that is `1 / 16 = 0.0625`, so **this
test cannot return `p < 0.05` no matter how strong the signal**. Five or
more groups are needed before significance is even attainable. Report
`p_min` alongside `p_value` so a non-significant result is not read as
evidence of absence.

Do not carry that floor across to the matrix test, whose blocked null
happens to have 16 points as well. The two floors differ by a factor of
two, and which way round is not arbitrary: swapping the trait labels of
*every* species reproduces the matrix test's statistic exactly, so its
maximum is always shared by at least two labellings and its floor is
`2 / 16 = 0.125`; the same global swap maps this test's annual-side
statistic onto its perennial-side one, which guarantees no tie at all, so
its floor stays `1 / 16 = 0.0625`. Neither test reaches 0.05 within
blocks. What the matrix test has instead is its free null over all 70
labellings, floor `2 / 70 = 0.029`; this one has only the within-pair
swap, since a free shuffle would draw designs that are not the paired one
the data came from.

It is run on the four within-genus contrasts, not on the all-pairs table:
the within-pair swap is what makes the null exact, and it needs the
paired design. `pair_name` has to name the contrasts as
`mod_results$classification` does, which since the all-pairs run above is
`"<sp1>.<sp2>"` rather than the genus label.

```{r module-parallel}
parallel_test <- tag_permutation(
  mod_results$classification, modules, orthologs,
  pairs = phylo_pairs |>
    transmute(
      sp1 = annual, sp2 = perennial,
      pair_name = paste(annual, perennial, sep = ".")
    ),
  group = trait_vec, target_group = "annual"
)

parallel_test$recurrence_table |>
  count(n_pairs, name = "n_hogs") |>
  show_table(caption = "HOG recurrence in modules diverged in the annual")

cat(
  "HOGs recurring in >=", parallel_test$min_recurrence, "pairs:",
  parallel_test$observed,
  "\n  p =", sprintf("%.4f", parallel_test$p_value),
  if (parallel_test$exact) "(exact)" else "(sampled)",
  "over",
  if (parallel_test$exact) {
    parallel_test$n_labellings
  } else {
    parallel_test$n_perm
  },
  "labellings from", parallel_test$n_swappable,
  "independent contrast group(s)",
  "\n  smallest attainable p:",
  sprintf("%.4f", max(parallel_test$p_min, parallel_test$p_attainable)),
  if (max(parallel_test$p_min, parallel_test$p_attainable) > 0.05) {
    "-- significance unreachable"
  } else {
    ""
  },
  "\n"
)
```

`min_recurrence` was left at its default, which is half the contrasts
able to supply a HOG, rounded up and floored at 2 --- so 2 here. A
constant would not describe a design of arbitrary size: at a per-side
rate of about 0.08, the chance a HOG reaches two of `k` sides unaided is
roughly 0.03 at four contrasts but 0.25 at twelve, so a fixed threshold
lets the statistic saturate on chance recurrence as contrasts are
added.

Two things to check before believing the result. The relabelling is only
exchangeable if the target side is not systematically the larger one, so
inspect `$pair_sizes` and `$size_asymmetry_p`; and because the statistic
is a raw count it responds strongly to how many HOGs the selected sides
hold. `statistic = "excess"` subtracts the count expected from
independent sides of those sizes, which helps when the sides are small
relative to the HOG universe and over-corrects when they are not --- see
`?tag_permutation` before relying on it.

```{r module-parallel-sizes}
show_table(parallel_test$pair_sizes,
  caption = "Diverged-HOG set sizes per contrast"
)

cat(
  "size-asymmetry sign test p:",
  sprintf("%.4f", parallel_test$size_asymmetry_p),
  "\n  (advisory at 0.10; like p_min it cannot reach 0.05 below five",
  "contrast groups)\n"
)
```

Running the same test for `target_group = "perennial"` reads a second
entry of the *same* null distribution --- relabelling every contrast maps
one trait's statistic onto the other's --- so the two are not independent
tests and must not be corrected as if they were.

### Hub genes and trait conservation

Identify hub genes within each species' modules, then classify each HOG
by whether it acts as a hub across trait groups. **Weighted degree** (the
default) is the recommended primary centrality for co-expression networks
-- it directly measures total co-expression strength to module neighbours,
which is the standard definition of a hub gene. All three centrality
measures (degree, betweenness, eigenvector) are reported regardless of
which is chosen as primary.

To disentangle centrality from regulatory function,
`characterize_hubs()` adds two metrics: **bridge fraction** (the
share of a hub's co-expression edges reaching genes in other
modules -- high values indicate a regulatory coordinator rather
than an intra-module effector) and **betweenness/degree ratio**
(importance for information flow relative to raw connectivity).
When an expression matrix is provided, it also computes the
**coefficient of variation** -- regulators tend to show higher
expression variability than their targets because they drive
transcriptional changes.

```{r module-hubs}
species_data <- species_data |>
  mutate(hubs = pmap(
    list(species, modules, network, reduced),
    \(sp, mod, net, red) {
      identify_module_hubs(mod, net, orthologs) |>
        characterize_hubs(mod, expr = red$expr_matrix)
    }
  ))

species_data |>
  transmute(species, trait,
    n_hubs  = map_int(hubs, \(h) sum(h$is_hub)),
    n_genes = map_int(hubs, nrow)
  ) |>
  show_table(caption = "Hub genes per species")
```

Deciding whether a HOG is a hub of *the same* module in two species is a
separate question from preservation, and `module_correspondence()`
answers it: it projects the resolved one-to-one ortholog map through both
partitions and tests each module pair with a hypergeometric. Because the
resolved map contributes one draw per test gene rather than one per
paralog, the test is no longer anti-conservative on multi-copy HOGs, and
its `jaccard` values run systematically higher than the retired
gene-overlap engine's -- the unchanged `jaccard_threshold = 0.1` inside
`classify_hub_conservation()` is therefore slightly more permissive than
it used to be. The q-values default to randomized-pi0 estimation and draw
from the global RNG, so the `conserved_hub` / `rewired_hub` verdicts are
seed-dependent; `set.seed()` first.

`classify_hub_conservation()` looks each species pair up under its
alphabetically sorted key, which is a real bug fixed here: the old raw
comparison list was keyed in the order the pairs were written
(`"HVUL.HJUB"`, `"VBRO.FPRA"`), so two of the four pairs never matched a
lookup and their hub calls silently degraded to `multi_trait_hub`.

```{r hub-conservation}
set.seed(42)

# Keyed on the ALPHABETICALLY SORTED pair, with module_sp1 belonging to
# the first of the two species -- that is what the lookup inside
# classify_hub_conservation() expects. preservation_paired() already ran
# the sorted direction and returns the map it resolved for it, so reuse
# that rather than resolving a second time.
correspondence <- list()
for (i in seq_len(nrow(phylo_pairs))) {
  sp <- sort(c(phylo_pairs$annual[i], phylo_pairs$perennial[i]))
  key <- paste(sp, collapse = ".")
  gene_map <- mod_results$raw[[key]]$map

  correspondence[[key]] <-
    module_correspondence(modules[[sp[1]]], modules[[sp[2]]], gene_map,
      sp_ref = sp[1], sp_test = sp[2]
    )
}

hub_class <- classify_hub_conservation(
  deframe(select(species_data, species, hubs)),
  trait_vec,
  module_comparisons = correspondence
)

hub_class |>
  filter(classification != "non_hub") |>
  count(classification) |>
  show_table(caption = "Hub conservation across all species")
```

The regulatory characterization helps separate potential regulators
from highly-connected effectors. We aggregate `bridge_fraction`,
`bt_degree_ratio`, and expression `cv` to the HOG level:

```{r hub-regulatory}
hub_regulatory <- species_data |>
  select(species, trait, hubs) |>
  unnest(hubs) |>
  filter(is_hub, !is.na(hog)) |>
  summarise(
    max_bridge = max(bridge_fraction, na.rm = TRUE),
    mean_bt_deg = mean(bt_degree_ratio, na.rm = TRUE),
    mean_cv = if (all(is.na(cv))) NA_real_ else mean(cv, na.rm = TRUE),
    .by = hog
  )

hub_class |>
  filter(classification != "non_hub") |>
  left_join(hub_regulatory, by = "hog") |>
  select(
    hog, classification, n_species_hub, max_centrality,
    max_bridge, mean_bt_deg, mean_cv
  ) |>
  arrange(desc(max_bridge), desc(mean_bt_deg)) |>
  show_table(
    caption = "Hub regulatory characterization (sorted by bridge fraction)"
  )
```

HOGs that are hubs specifically in annual (or perennial) species -- the
`annual_specific_hub`, `perennial_specific_hub`, and `rewired_hub`
classes -- are candidates for parallel regulatory evolution. Ranking
by `bridge_fraction` and `bt_degree_ratio` prioritizes those most
likely to be true regulators:

```{r hub-trait-specific}
hub_class |>
  filter(classification %in% c(
    "annual_specific_hub",
    "perennial_specific_hub",
    "rewired_hub"
  )) |>
  left_join(hub_regulatory, by = "hog") |>
  select(
    hog, classification, n_species_hub, hub_trait_groups,
    max_centrality, max_bridge, mean_bt_deg, mean_cv
  ) |>
  arrange(desc(max_bridge)) |>
  show_table(
    caption = "Trait-specific and rewired hubs (ranked by regulatory potential)"
  )
```

### Co-expression partners of hub HOGs

After identifying trait-specific hubs, the natural question is: "Which
other HOGs co-express with this hub across species?"
`get_coexpressed_hogs()` resolves a HOG to its gene(s) in each species,
extracts co-expression neighbors from the network, maps them back to
HOGs, and aggregates across species. The `coexpressed_traits` column
makes trait-group filtering straightforward:

```{r hub-partners}
candidate <- hub_class |>
  filter(classification %in% c(
    "annual_specific_hub",
    "perennial_specific_hub",
    "rewired_hub", "conserved_hub"
  )) |>
  slice_max(n_species_hub, n = 1, with_ties = FALSE) |>
  pull(hog)

if (length(candidate) == 1) {
  partners <- get_coexpressed_hogs(
    candidate, networks, orthologs,
    species_trait = trait_vec, min_species = 2L,
    edges = edges
  )

  cat(
    candidate, ":", nrow(partners), "partner HOGs in >= 2 species —",
    sum(partners$coexpressed_traits == "annual", na.rm = TRUE), "annual-only,",
    sum(grepl(",", partners$coexpressed_traits), na.rm = TRUE), "cross-trait\n"
  )

  partners |>
    slice_head(n = 20) |>
    show_table(caption = paste("Top co-expression partners of", candidate))
} else {
  cat("No trait-specific or conserved hubs found\n")
}
```

## 4. Edge robustness: density sweep

**Network density** controls how many edges are retained in the
co-expression network. At 3% density, only the top 3% of all
pairwise MR values pass the threshold -- these are the strongest
co-expression relationships.

Networks keep edges at or *above* the threshold, so multipliers above
1 tighten the network and multipliers below 1 loosen it. Raw MR
values are tightly compressed near the top of the distribution, so
small multipliers move the effective density a lot: on this dataset
1.025x already cuts the density to ~1.5% and 0.975x loosens it to
~4.7%. We sweep 11 levels from 0.975x to 1.025x the baseline
threshold, corresponding to effective densities of approximately
4.7%--1.5% around the 3% midpoint:

- **Tighter** (1.005x--1.025x, effective density ~2.7%--1.5%): does
  the signal survive substantial edge removal? Surviving
  co-expressologs are strong, not marginal.
- **Looser** (0.995x--0.975x, effective density ~3.3%--4.7%): is the
  signal specific to this density range, or does it persist as new
  edges are added? Cliques that only appear at very loose thresholds
  are likely noise from extra edges.

The 3% density choice is somewhat arbitrary. Sweeping a range around
it reveals sensitivity to that choice -- a truly robust
co-expressolog should be present across a range of densities, not
just at one particular cutoff.

Since networks are stored sparsely (v0.2.0), the loosest multiplier
must stay within the stored edge superset, and the sweep errors if a
multiplier falls below the store. The networks in section 1 were
built with `store_density = 0.06`, so the loosest level (0.975x,
~4.7%) keeps a comfortable margin to the store (a 6% store
corresponds to a ~0.96x multiplier on this dataset; the default 5%
store would sit at ~0.975x, i.e. exactly at the loosest level).

The permutation pipeline is re-run at each level because network
neighborhoods change with the threshold, altering the HOG-level
q-values. The floor is derived from the networks themselves rather
than assumed from the prose above, so the chunk keeps working if the
expression matrices or `reduce_orthogroups()` output shift the raw-MR
quantiles:

```{r density-sweep}
multipliers <- seq(0.975, 1.025, by = 0.005)
store_floor <- max(vapply(
  networks, \(n) n$store_threshold / n$threshold,
  numeric(1)
))
stopifnot(min(multipliers) >= store_floor)

sweep <- density_sweep(networks, orthologs,
  multipliers = multipliers,
  method = "permutation", use_torch = TRUE, n_cores = n_cores
)

# Edge survival across density levels
sweep |>
  transmute(
    Multiplier = paste0(multiplier, "x"),
    Eff_density = sprintf("%.2f%%", eff_density * 100),
    Conserved_edges = n_significant,
    Baseline = if_else(multiplier == 1, "***", "")
  ) |>
  show_table(caption = "Conserved edges across density levels")
```

The mean effective density masks species-level variation. Because
each species has a different MR distribution shape, a uniform
threshold multiplier produces different effective densities across
species:

```{r species-density-variation}
baseline_densities <- sweep |>
  filter(multiplier == 1) |>
  pull(species_densities) |>
  pluck(1)

tibble(
  species = names(baseline_densities),
  trait = trait_vec[names(baseline_densities)],
  eff_density = sprintf("%.2f%%", baseline_densities * 100)
) |>
  show_table(caption = "Per-species effective density at baseline (1.0x)")
```

The edge count may not decrease monotonically because the permutation
test operates at the HOG level: tighter neighborhoods can sharpen the
signal for some HOGs while eliminating it for others. Local fluctuations
reflect the discrete nature of the permutation q-values. The
interpretation is asymmetric: surviving tighter thresholds indicates
strong signal, while surviving looser thresholds indicates specificity
(not merely noise from additional edges).

**Analytical alternative: `clique_persistence()`.**
The density sweep above is rigorous but expensive — it re-runs the
full permutation pipeline at each threshold level. For a quick estimate,
`clique_persistence()` computes the ratio of each co-expressolog edge's
MR value to the species' density threshold. A persistence of 1.5 means
the weakest supporting edge is 50% above threshold and would survive a
1.5x tightening. This is a principled analytical proxy: it correctly
identifies the weakest link in each clique's co-expressolog support.
However, it cannot detect that tighter neighborhoods may change which
HOGs reach significance in the permutation test, or that new cliques
may form from HOGs that become significant at stricter thresholds.
Use persistence for screening or when GPU is unavailable; use the full
sweep for publication-quality robustness assessment.

## 5. Clique detection and structural stability

**Phylogenetic structure of pairwise comparisons.** The
co-expressolog edge set contains C(8,2) = 28 species pairs, mixing
three types: (1) **focal pairs** from the same genus (4 annual vs
perennial sister pairs), (2) **within-trait pairs** comparing two
annuals or two perennials from different genera (12 pairs), and (3)
**cross-trait/cross-genus pairs** (12 pairs). Cliques built on all
28 pairs can span phylogenetically non-independent comparisons.
`classify_cliques()` handles this by separating within-trait from
cross-trait cliques in the waterfall, but cliques are not weighted
by phylogenetic distance. For the current dataset this is acceptable:
each genus contributes exactly one species per trait, limiting
phylogenetic confounding.

**A second backend.** `find_cliques()`, used below, builds cliques of a
per-orthogroup *species* graph and returns one best gene assignment each.
`gene_clique_graph()` builds maximal cliques of the per-orthogroup *gene*
graph instead, following Rodriguez *et al.* (2026,
doi:10.1038/s41467-026-75624-2), so a multi-copy HOG can yield several
overlapping cliques, and `classify_gene_cliques()` applies that paper's
five-tier taxonomy. Its two tolerance tiers cover different kinds of gap:
`partial_significant` admits an edge to the graph at a loose threshold but
counts it as evidence only at the strict one (weak wiring), while
`partial_present` accepts a fully significant clique that is one species
short (a missing gene). Section 6 runs both backends: the species-graph
waterfall first, then the gene-graph taxonomy on the same data, with
guidance on which one answers which question.

Using edges from the baseline density level, we find cliques and
test their structural stability via leave-k-out jackknife. The
jackknife removes k species at a time and checks whether each
clique's gene assignments are preserved (purely structural,
trait-agnostic). Trait annotations are added post-hoc:

```{r cliques-stability}
# Use the baseline (1.0x) edges for clique detection and stability
robust_edges <- sweep |>
  filter(multiplier == 1) |>
  pull(edges) |>
  pluck(1)
cat(
  "Using", sum(robust_edges$type == "conserved"), "conserved edges at",
  sprintf(
    "%.2f%%",
    (sweep |> filter(multiplier == 1) |> pull(eff_density)) * 100
  ),
  "density\n\n"
)

# Find cliques
cliques <- find_cliques(robust_edges, species, min_species = 2L)
cat(nrow(cliques), "cliques\n")

cliques |>
  count(n_species) |>
  show_table(pageLength = 10, caption = "Cliques by number of species")
```

Cliques with `n_species = 2` are single conserved edges between one
pair of species -- the weakest form of co-expression conservation.
The most compelling evidence comes from cliques spanning 3 or more
species, as these require coordinated conservation across multiple
independent lineages.

The conserved edges feeding into clique detection come from three
types of species pair:

```{r pair-type-breakdown}
focal_set <- paste(phylo_pairs$annual, phylo_pairs$perennial)
robust_edges |>
  filter(type == "conserved") |>
  mutate(
    t1 = trait_vec[species1], t2 = trait_vec[species2],
    pair_type = case_when(
      paste(species1, species2) %in% focal_set |
        paste(species2, species1) %in% focal_set ~ "focal (sister pair)",
      t1 == t2 ~ "within-trait",
      TRUE ~ "cross-trait/cross-genus"
    )
  ) |>
  count(pair_type) |>
  show_table(caption = "Conserved edges by pair type")
```

```{r stability}
# Jackknife stability (all k from 1 to N-2)
max_k <- length(species) - 2L
stab <- clique_stability(robust_edges, species,
  species_trait = trait_vec,
  all_species = species,
  full_cliques = cliques,
  max_k = max_k
)
cat("Cliques tested:", length(stab$stability_class), "\n")
cat("Structurally stable at k=1:", sum(stab$stability_class >= 1L), "\n")

tibble(stability_class = stab$stability_class) |>
  mutate(stability_class = factor(stability_class, levels = 0:max_k)) |>
  count(stability_class, .drop = FALSE) |>
  show_table(pageLength = 10, caption = "Stability class distribution")
```

## 6. HOG classification with robustness ranking

Classify each HOG by its conservation pattern, then rank within each
category by robustness. The waterfall assigns each HOG to exactly one
class: complete > partial > differentiated > trait_specific > unclassified.

```{r classify}
result <- classify_cliques(robust_edges, species, trait_vec,
  min_species = 2L,
  stability = stab
)

result |>
  count(classification) |>
  show_table(pageLength = 5, caption = "HOG classification summary")
```

Restricting to cliques spanning 3 or more species (excluding single-edge
2-species signals) gives a clearer picture of multi-lineage conservation:

```{r focal-classification}
result |>
  filter(n_species >= 3 | is.na(n_species)) |>
  count(classification) |>
  show_table(
    pageLength = 5,
    caption = "HOG classification (cliques >= 3 species only)"
  )
```

- **complete**: all 8 species form a clique -- deeply conserved across
  all Pooideae
- **partial**: clique spans both annual and perennial species, but not
  all 8
- **differentiated**: annual species have a clique AND perennial species
  have a clique, but no cross-group conservation -- lineage-specific
  co-expression programs
- **trait_specific**: only annuals or only perennials have a clique
- **unclassified**: no conservation signal detected

The robustness ranking comes from combining density sweep survival
with stability class. HOGs classified at a stricter density level
with high stability class are the strongest findings:

```{r ranking}
# Count how many density levels each HOG's edges survive
hog_survival <- sweep |>
  transmute(multiplier,
    hog = map(edges, \(e) {
      e |>
        filter(type == "conserved") |>
        distinct(hog) |>
        pull(hog)
    })
  ) |>
  unnest(hog) |>
  count(hog, name = "n_densities_survived")

result <- result |>
  left_join(hog_survival, by = "hog") |>
  mutate(n_densities_survived = replace_na(n_densities_survived, 0L))

# Annotate trait composition from the best clique per HOG
hog_traits <- cliques |>
  group_by(hog) |>
  slice_min(mean_q, n = 1, with_ties = FALSE) |>
  ungroup() |>
  pivot_longer(all_of(species), names_to = "sp", values_to = "gene") |>
  filter(!is.na(gene)) |>
  summarise(
    n_annual = sum(trait_vec[sp] == "annual"),
    n_perennial = sum(trait_vec[sp] == "perennial"),
    .by = hog
  )

result <- left_join(result, hog_traits, by = "hog")

# Show ranked HOGs per category
show_cols <- c(
  "hog", "n_species", "n_annual", "n_perennial",
  "best_mean_q", "stability_class", "n_densities_survived"
)
show_cols <- intersect(show_cols, names(result))
```

Every clique now has a `stability_class` -- the highest k at which it
survives ALL species-removal subsets. This is a purely structural measure.
Trait annotations (from the `traits` column in the stability output) are
orthogonal: they describe which life-cycle groups each clique spans, not
whether the clique is robust.

### {.tabset}

#### Partial

```{r partial-table}
show_classification(
  result, "partial", show_cols,
  "Partial HOGs (cross-trait cliques)"
)
```

#### Trait-specific

```{r trait-specific-table}
show_classification(
  result, "trait_specific", show_cols,
  "Trait-specific HOGs (one lineage only)"
)
```

#### Differentiated

```{r differentiated-table}
show_classification(
  result, "differentiated", show_cols,
  "Differentiated HOGs (within-group conservation, cross-group divergence)"
)
```

#### Complete

```{r complete-table}
show_classification(
  result, "complete", show_cols,
  "Complete HOGs (all 8 species)"
)
```

###

### The gene-graph taxonomy: which classifier to reach for

`classify_cliques()` above and `classify_gene_cliques()` below answer
different questions, and neither is deprecated in favour of the other.

`classify_cliques()` works on the *species* graph. It asks, per HOG,
which species are joined by conserved co-expression and whether that
pattern respects the trait split, and it returns one row per HOG.
Because `find_cliques()` commits to one best gene assignment per
species clique, a multi-copy HOG still gets a single answer, and that
answer is what `clique_stability()` and `clique_threshold_sweep()`
consume -- the robustness ranking earlier in this section only exists
on this side.

`classify_gene_cliques()` works on the *gene* graph built by
`gene_clique_graph()`. It asks which individual gene copies are
mutually conserved, so one HOG can produce several overlapping cliques
and the answer names paralogs rather than species. It applies the
five-tier taxonomy of Rodriguez *et al.* (2026) with two explicit
tolerance tiers, and its lineage split is an argument rather than the
trait vector, so it can be run against a clade partition the trait does
not follow. Reach for it when which copy sits in the conserved core
matters, or when the published taxonomy is what you need to report;
reach for `classify_cliques()` for a one-row-per-HOG trait summary
wired into the stability and sweep machinery.

The gene graph below is built from `edges_analytical`, not from the
permutation `edges`. The tolerance tiers need q-values that reach up to
`alpha_graph`, and the Besag--Clifford permutation q-values on this
dataset stop just above 0.4 -- adaptive stopping never lets them climb
higher -- so a graph at 0.9 would be the *complete* ortholog graph.
That is not a corner case: enumerating it here produces roughly 13.7
million clique-member rows and leaves `partial_significant` with
nothing to relax against. The analytical q-values run up to 0.98, so
the 0.9 graph is a genuine relaxation of the 0.1 calling threshold:

```{r gene-cliques}
gene_cliques <- gene_clique_graph(edges_analytical,
  min_size = 3L,
  alpha_graph = 0.9
)

gene_tiers <- classify_gene_cliques(gene_cliques, edges_analytical,
  species,
  lineage = trait_vec,
  alpha_call = 0.1, alpha_graph = 0.9
)

gene_tiers |>
  count(classification) |>
  show_table(pageLength = 6, caption = "Gene-clique tiers")
```

`alpha_call = 0.1` is the function default and is looser than the
`alpha = 0.05` used to call edges conserved in section 2, so these
tiers are not a subset of the `type == "conserved"` rows.

One permissive run is enough on this dataset: every clique the 0.1
graph produces is re-found in the 0.9 graph, so row-binding a strict
run only duplicates it (`complete_conserved` goes from 391 to exactly
782). That is not guaranteed in general -- a clique maximal at the
calling threshold can be absorbed into a larger one at the loose
threshold -- so combine both runs, each with its own `id_prefix`, when
the strict-graph cliques matter in their own right.

`lineage_specific` and `differentiated` come back empty under
`lineage = trait_vec`, and both absences are informative rather than a
missing argument. `lineage_specific` would need a clique covering one
entire life-cycle group with every species outside it *untestable*
against the clique; a species that was compared and came back
non-significant blocks the tier, and here every species pair is tested.
`differentiated` would need each life-cycle group fully significant
within itself while at most `cross_max` = 9 of the 16 cross-trait pairs
are significant, which no clique on this dataset satisfies.

`hog_class` collapses the per-clique tiers back to one row per HOG by
the same precedence, which is the level the published scripts report:

```{r gene-cliques-hog}
gene_tiers |>
  distinct(hog, hog_class) |>
  count(hog_class) |>
  show_table(pageLength = 6, caption = "Gene-clique tiers per HOG")
```

### Clique robustness diagnostics

The density sweep and jackknife stability above assess whether cliques
survive threshold changes and species removal. Several additional
diagnostics quantify clique quality at a finer grain.

**Edge-level quality metrics.** `find_cliques()` now returns three
columns alongside each clique: `intensity` (Onnela intensity -- the
geometric mean, across present edges, of each edge's effect-size
percentile among all tested pairs of its species pair), `coherence`
(intensity divided by the arithmetic mean, equal to 1 when all edge
weights are equal), and `min_effect_size` (the bottleneck enrichment
across present edges). The weights are percentiles rather than
`1 - q.value` because every clique edge has already passed alpha, so
`1 - q` would sit near 1 for every clique; this is also why
`find_cliques()` should receive the unfiltered edge table. High intensity with high coherence indicates
uniformly strong conservation; high intensity with low coherence
flags cliques carried by one or two dominant edges. The
`cost_weights` parameter controls gene-assignment ranking: the
default `c(q = 1, effect = 0)` minimises mean q-value, while
`c(q = 1, effect = 1)` uses a composite cost
`q * mean_q - effect * mean_effect` that also rewards large
effect sizes:

```{r intensity-coherence, eval = FALSE}
# Intensity and coherence are already in cliques from find_cliques()
cliques |>
  select(hog, n_species, mean_q, intensity, coherence, min_effect_size) |>
  filter(n_species >= 3) |>
  arrange(desc(intensity)) |>
  head(20) |>
  show_table(caption = "Top cliques by Onnela intensity (>= 3 species)")

# Composite cost: rank gene assignments by both q-value and effect size
cliques_composite <- find_cliques(robust_edges, species,
  min_species = 2L,
  cost_weights = c(q = 1, effect = 0.5)
)
```

**Jaccard overlap in edge output.** `comparison_to_edges()` (and
therefore `find_coexpressologs()`) now includes a `jaccard` column:
the geometric mean of the two directional Jaccard indices of
neighborhood overlap. This complements the hypergeometric q-value
with a non-parametric measure of neighborhood similarity (range
0--1). Large effect size with low Jaccard suggests enrichment driven
by a small shared subset rather than broad neighborhood overlap:

```{r edge-jaccard, eval = FALSE}
robust_edges |>
  filter(type == "conserved") |>
  select(hog, gene1, gene2, species1, species2, q.value, effect_size, jaccard) |>
  arrange(desc(jaccard)) |>
  head(20) |>
  show_table(caption = "Top conserved edges by Jaccard overlap")
```

**Birth/death persistence.** `clique_threshold_sweep()` returns a
`$persistence` element with one row per baseline clique. Each row
records the `birth` (lowest multiplier where the clique exists),
`death` (first multiplier above birth where it disappears), and
`persistence` (death minus birth). A persistence of `NA` means the
clique survived all tested multipliers. Unlike the density sweep
survival count, persistence captures the continuous range of
thresholds over which each clique lives:

```{r birth-death, eval = FALSE}
# clique_threshold_sweep() scales the same baseline threshold as the
# density sweep above (raw MR, tightly compressed near the top of the
# distribution), so its multipliers live in the same near-1 band --
# c(1.5, 2, 3, ...) would push every level past the empty-network
# threshold this dataset hits above ~1.03x.
sweep_result <- clique_threshold_sweep(
  cliques, species, networks, orthologs,
  multipliers = c(1.005, 1.01, 1.015, 1.02, 1.025),
  min_species = 2L, n_cores = n_cores
)

sweep_result$persistence |>
  arrange(desc(persistence)) |>
  head(20) |>
  show_table(caption = "Clique birth/death persistence")
```

**Perturbation test.** `clique_perturbation_test()` adds Gaussian
noise to MR scores in each species network, re-thresholds, re-runs
the comparison-to-cliques pipeline, and measures how often each
baseline clique survives (Jaccard >= threshold to its perturbed
counterpart). This tests robustness to measurement noise without
recomputing correlations from scratch. Cliques with high
`survival_rate` are insensitive to small perturbations in edge
weights:

```{r perturbation, eval = FALSE}
pert <- clique_perturbation_test(
  cliques, species, networks, orthologs,
  n_boot = 100, noise_sd = 0.1,
  min_species = 2L, jaccard_threshold = 0.5,
  n_cores = n_cores, seed = 42
)

pert |>
  arrange(desc(survival_rate)) |>
  head(20) |>
  show_table(caption = "Clique perturbation survival (100 bootstrap iterations)")
```

**Intensity Z-score.** `clique_intensity_test()` builds a null
distribution by permuting the ortholog mapping (globally shuffling
Species2 genes across all rows, destroying both the specific mapping
and within-HOG gene grouping). For each permutation, cliques are
re-detected and the intensity of the best-matching clique (by
Jaccard) is recorded. The Z-score and empirical p-value test whether
the observed gene-to-gene correspondence produces stronger
co-expression conservation than random mappings:

```{r intensity-test, eval = FALSE}
intens <- clique_intensity_test(
  cliques, species, networks, orthologs,
  n_perm = 500, min_species = 2L,
  n_cores = n_cores, seed = 42, edges = robust_edges
)

intens |>
  filter(p_value < 0.05) |>
  arrange(desc(z_score)) |>
  head(20) |>
  show_table(caption = "Cliques with significantly elevated intensity (p < 0.05)")
```

Together, these diagnostics address complementary robustness
questions: density sweep tests threshold sensitivity, jackknife
tests species-removal stability, perturbation tests measurement
noise, and the intensity Z-score tests whether the ortholog mapping
itself drives the signal. Cliques passing all four are the strongest
candidates for biologically meaningful co-expression conservation.

## 7. Biological interpretation

The central question is whether independent annual Pooideae lineages
draw on a shared, latent regulatory architecture inherited from
perennial ancestors. The analyses above provide several lines of
evidence:

```{r synthesis}
n_complete <- sum(result$classification == "complete", na.rm = TRUE)
n_partial <- sum(result$classification == "partial", na.rm = TRUE)
n_differentiated <- sum(result$classification == "differentiated", na.rm = TRUE)
n_trait_specific <- sum(result$classification == "trait_specific", na.rm = TRUE)

# Robust candidates: survive >= half the density levels and stability >= 1
n_robust <- sum(
  result$n_densities_survived >= ceiling(nrow(sweep) / 2) &
    !is.na(result$stability_class) & result$stability_class >= 1L,
  na.rm = TRUE
)

cat("Complete (all 8 species):", n_complete, "\n")
cat("Partial (cross-trait, < 8):", n_partial, "\n")
cat(
  "Differentiated (independent annual + perennial cliques):",
  n_differentiated, "\n"
)
cat("Trait-specific (one lineage only):", n_trait_specific, "\n")
cat(
  "Robust candidates (>= 50% density survival + stability >= 1):",
  n_robust, "\n"
)
if (exists("parallel_test")) {
  cat("\nParallel module signal: ", parallel_test$observed,
    " HOGs recur in modules diverged in the annual across >= 2 pairs (p = ",
    sprintf("%.4f", parallel_test$p_value), ", floor ",
    sprintf("%.4f", max(
      parallel_test$p_min,
      parallel_test$p_attainable
    )), ")\n",
    sep = ""
  )
}
```

**Differentiated HOGs** are the strongest candidates for regulatory
divergence: the same ortholog group has co-expression relationships in
both annuals and perennials, but these programs are incompatible (no
cross-group clique). This pattern is consistent with a shared ancestral
regulatory architecture that has been independently remodelled in each
annual lineage.

**Trait-specific HOGs** (annual-only or perennial-only cliques) may
encode lineage-specific regulatory modules. The formal permutation
test above addresses whether modules diverged in the annual share HOGs
across independent lineages more than expected by chance. A significant
result supports the hypothesis that annuality draws on a shared
regulatory toolkit rather than arising from independent *de novo*
rewiring in each lineage.

**Robustness ranking** combines density sweep survival with jackknife
stability class. HOGs classified at tighter density levels with high
stability represent the strongest findings for experimental follow-up.

**Regulatory characterization** helps prioritize among hub
candidates. Hubs with high `bridge_fraction` connect across modules,
a hallmark of regulatory coordinators (TFs, signalling genes). High
`bt_degree_ratio` indicates importance for information flow relative
to raw connectivity. High expression `cv` suggests the gene drives
transcriptional changes rather than passively responding. Candidates
scoring high on all three metrics are the strongest regulatory hub
candidates, especially when they also carry known DNA-binding or
signalling domains (testable via an annotation overlay with
PlantTFDB or InterPro).

**Caveats.** Co-expression conservation is necessary but not sufficient
evidence for regulatory conservation. Shared co-expression could
reflect shared ancestry (phylogenetic inertia) rather than convergent
regulatory function. The paired phylogenetic design partially controls
for this -- each pair is an independent comparison within a genus --
but formal ancestral state reconstruction of co-expression networks
would be needed to fully distinguish these scenarios. The candidate
HOGs identified here should be validated with functional annotation
(GO enrichment, transcription factor binding motifs) and, where
available, tissue- or stage-specific expression data.

## Session info

```{r sessioninfo}
sessionInfo()
```
