---
title: "Comparative co-expression analysis with rcomplex"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparative co-expression analysis with rcomplex}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_se <- requireNamespace("SummarizedExperiment", quietly = TRUE)
has_torch <- requireNamespace("torch", quietly = TRUE) &&
             torch::torch_is_installed() &&
             tryCatch({ torch::torch_randn(1); TRUE }, error = function(e) FALSE)
has_tidy <- requireNamespace("purrr", quietly = TRUE) &&
            requireNamespace("tibble", quietly = TRUE) &&
            requireNamespace("stringr", quietly = TRUE)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      eval = has_se && has_torch && has_tidy)

# Helper for interactive tables (sortable by one or multiple columns)
show_table <- function(df, pageLength = 10, ...) {
  if (requireNamespace("DT", quietly = TRUE)) {
    DT::datatable(df, rownames = FALSE,
                  options = list(pageLength = pageLength, scrollX = TRUE,
                                 ordering = TRUE, orderMulti = TRUE),
                  ...)
  } else {
    knitr::kable(df)
  }
}

# Helper for ranked classification tables
show_classification <- function(df, class, cols, caption) {
  sub <- df |>
    dplyr::filter(classification == class) |>
    dplyr::select(dplyr::any_of(cols)) |>
    dplyr::arrange(dplyr::desc(n_densities_survived), best_mean_q)
  show_table(sub, pageLength = 10, caption = caption)
}
```

This vignette demonstrates a complete rcomplex analysis using expression
data from eight Pooideae grass species spanning four phylogenetic pairs,
each containing one annual and one perennial species:

| Species | Code | Life cycle | Pair |
|---------|------|------------|------|
| *Brachypodium distachyon* | BDIS | Annual | Brachypodium |
| *Brachypodium sylvaticum* | BSYL | Perennial | Brachypodium |
| *Hordeum vulgare* | HVUL | Annual | Hordeum |
| *Hordeum jubatum* | HJUB | Perennial | Hordeum |
| *Briza maxima* | BMAX | Annual | Briza |
| *Briza media* (crossmapped to BMAX) | BMED | Perennial | Briza |
| *Vulpia bromoides* | VBRO | Annual | Festuca/Vulpia |
| *Festuca pratensis* | FPRA | Perennial | Festuca/Vulpia |

The dataset contains the 1500 most highly expressed HOGs from VST-normalized RNA-seq from the
"Live or let die?" project, stored as SummarizedExperiment objects.

## 1. Data preparation

```{r data, message = FALSE}
library(rcomplex)
library(SummarizedExperiment)
library(dplyr, warn.conflicts = FALSE)
library(tidyr)
library(purrr)
library(tibble)

se_list <- readRDS(system.file("extdata", "pooideae_vignette.rds",
                                package = "rcomplex"))
```

We organise all per-species data in a single nested tibble. Each column
accumulates one processing stage — the `tribble` defines species metadata,
then `mutate` + `map` appends the SE objects, paralog reduction, and
network construction as list-columns:

### Paralog reduction

Multi-copy gene families often contain recent duplicates with nearly
identical expression profiles. These inflate the combinatorial search
space for clique detection without adding information.
`reduce_orthogroups()` merges correlated paralogs within each HOG
using Ward.D2 hierarchical clustering on Pearson correlation distance.
Paralogs above the correlation threshold (default 0.7) are replaced
by their averaged expression; subfunctionalized copies with distinct
expression programs are preserved:

```{r species-data}
species_data <- tribble(
  ~species, ~trait,
  "BDIS",   "annual",
  "BSYL",   "perennial",
  "HVUL",   "annual",
  "HJUB",   "perennial",
  "BMAX",   "annual",
  "BMED",   "perennial",
  "VBRO",   "annual",
  "FPRA",   "perennial"
) |>
  mutate(
    se = se_list[species],
    reduced = map(se, \(s) {
      expr <- SummarizedExperiment::assay(s, 1)
      rd <- SummarizedExperiment::rowData(s)
      ortho_sp <- tibble(gene = rownames(s),
                         hog = as.character(rd$hog)) |>
        filter(!is.na(hog))
      reduce_orthogroups(expr, ortho_sp, gene_col = "gene",
                         cor_threshold = 0.7)
    })
  )

species_data |>
  select(species, reduced) |>
  pwalk(\(species, reduced)
    cat(species, ":", reduced$n_original, "->", reduced$n_reduced, "genes",
        "(", reduced$n_merged, "merged )\n"))
```

### Network construction

Build co-expression networks from the reduced expression matrices
using Mutual Rank (MR) normalization at 3% density:

```{r networks}
species_data <- species_data |>
  mutate(network = map(reduced, \(r)
    compute_network(r$expr_matrix, norm_method = "MR", density = 0.03)))

# Named lists for functions that expect them
species  <- pull(species_data, species)
trait_vec <- deframe(select(species_data, species, trait))
networks <- deframe(select(species_data, species, network))
```

### Ortholog extraction

Extract ortholog pairs from HOG annotations in rowData. After
paralog reduction, gene names may have changed (merged representatives),
so we map through each species' reduction via joins. Gene IDs are
species-unique (genome-specific identifiers), so each species' map
is applied only to the column containing that species' genes:

```{r orthologs}
# All species pairs as a tibble
pairs <- combn(species, 2, simplify = FALSE) |>
  map(\(p) tibble(sp1 = p[1], sp2 = p[2])) |>
  list_rbind()

# Extract orthologs for every pair, then map gene names through reduction
gene_maps <- species_data |>
  transmute(species, gm = map(reduced, "gene_map")) |>
  unnest(gm) |>
  select(species, original, representative)

orthologs <- pairs |>
  mutate(data = map2(sp1, sp2, \(a, b)
    extract_orthologs(se_list[[a]], se_list[[b]]))) |>
  unnest(data) |>
  # Map Species1 genes via sp1's reduction
  left_join(gene_maps |> select(sp1 = species,
                                 Species1 = original, rep1 = representative),
            by = c("sp1", "Species1")) |>
  mutate(Species1 = coalesce(rep1, Species1)) |>
  # Map Species2 genes via sp2's reduction
  left_join(gene_maps |> select(sp2 = species,
                                 Species2 = original, rep2 = representative),
            by = c("sp2", "Species2")) |>
  mutate(Species2 = coalesce(rep2, Species2)) |>
  select(-sp1, -sp2, -rep1, -rep2) |>
  distinct()

cat(nrow(orthologs), "ortholog pairs across",
    nrow(pairs), "species pairs\n")
```

## 2. Pairwise comparison and HOG-level testing

For each of the C(8,2) = 28 species pairs, we run two analyses:

1. **Neighborhood comparison** (`compare_neighborhoods()`): hypergeometric
   test for overlap of co-expression neighborhoods, producing per-pair
   p-values and fold-enrichment effect sizes.
2. **HOG-level permutation test** (`permutation_hog_test()`): gene-identity
   permutation within each HOG. This is necessary because pair-level tests
   within a HOG are correlated (shared neighborhoods), making Fisher's
   method or Storey q-values anti-conservative. The permutation p-value is
   exact regardless of the dependency structure.

With `use_torch = TRUE`, the fold-enrichment matrix is precomputed on
GPU, reducing each permutation to fast table lookups.

The HOG-level q-values determine which edges are co-expressologs
(conserved co-expression relationships):

```{r comparison}
# Helper: build edges from raw comparison + HOG-level q-values
build_perm_edges <- function(comp, hog_res, sp_a, sp_b, alpha = 0.05) {
  comp |>
    as_tibble() |>
    left_join(hog_res |> select(hog, q.value), by = "hog") |>
    transmute(
      gene1 = Species1,
      gene2 = Species2,
      species1 = sp_a,
      species2 = sp_b,
      hog,
      q.value,
      effect_size = sqrt(Species1.effect.size * Species2.effect.size),
      type = if_else(!is.na(q.value) & q.value < alpha, "conserved", "ns")
    )
}

n_cores <- min(4L, parallel::detectCores(logical = FALSE))

# Run all 28 pairs: nested tibble with comp + hog_res as list-columns
comparisons <- pairs |>
  mutate(
    comp = map2(sp1, sp2, \(a, b)
      compare_neighborhoods(networks[[a]], networks[[b]], orthologs)),
    hog_res = pmap(list(sp1, sp2, comp), \(a, b, c) {
      if (nrow(c) == 0) return(NULL)
      permutation_hog_test(networks[[a]], networks[[b]], c,
                           n_cores = n_cores, use_torch = TRUE)
    })
  ) |>
  filter(map_int(comp, nrow) > 0)

edges <- comparisons |>
  pmap(\(sp1, sp2, comp, hog_res, ...)
    build_perm_edges(comp, hog_res, sp1, sp2)) |>
  list_rbind()

all_hog <- comparisons |>
  transmute(hog_res = map2(hog_res, paste(sp1, sp2, sep = "-"), \(h, pair)
    h |> as_tibble() |> mutate(pair = pair))) |>
  pull(hog_res) |>
  list_rbind()

cat(nrow(edges), "edges:", sum(edges$type == "conserved"), "conserved,",
    sum(edges$type == "ns"), "non-significant\n")
cat(nrow(all_hog), "HOG tests,",
    sum(all_hog$q.value < 0.05), "significant (q < 0.05)\n")
```

For quick exploration without GPU, `run_pairwise_comparisons()` provides
a faster analytical path using pair-level Storey q-values (less
statistically rigorous for multi-copy HOGs):

```{r analytical-edges}
edges_analytical <- run_pairwise_comparisons(networks, orthologs)
cat("Analytical:", sum(edges_analytical$type == "conserved"), "conserved edges\n")
cat("Permutation:", sum(edges$type == "conserved"), "conserved edges\n")
```

## 4. Module-level analysis

Detect co-expression modules with Leiden community detection, then
compare across species:

```{r modules}
mod_bdis <- detect_modules(networks$BDIS, method = "leiden", resolution = 1.0)
mod_bsyl <- detect_modules(networks$BSYL, method = "leiden", resolution = 1.0)

mod_comp <- compare_modules(mod_bdis, mod_bsyl, orthologs,
                             method = "hypergeometric")
classes <- classify_modules(mod_comp)

classes |>
  count(classification) |>
  show_table(caption = "Module classification (BDIS vs BSYL)")
```

### Hub genes and trait conservation

Identify hub genes within each species' modules, then classify each HOG
by whether it acts as a hub across trait groups. **Weighted degree** (the
default) is the recommended primary centrality for co-expression networks
-- it directly measures total co-expression strength to module neighbours,
which is the standard definition of a hub gene. All three centrality
measures (degree, betweenness, eigenvector) are reported regardless of
which is chosen as primary:

```{r module-hubs}
# Comparison result for BDIS vs BSYL (for conservation tie-breaking)
comp_bdis_bsyl <- comparisons |>
  filter(sp1 == "BDIS", sp2 == "BSYL") |>
  pull(comp) |>
  pluck(1)

# Identify hubs (top 10% per module by weighted degree)
hubs_bdis <- identify_module_hubs(mod_bdis, networks$BDIS, orthologs,
                                   comparison = comp_bdis_bsyl)
hubs_bsyl <- identify_module_hubs(mod_bsyl, networks$BSYL, orthologs,
                                   comparison = comp_bdis_bsyl)

show_table(hubs_bdis |> filter(is_hub),
           caption = "Hub genes in BDIS modules")
```

```{r hub-conservation}
# Classify hub conservation across traits
hub_trait <- c(BDIS = "annual", BSYL = "perennial")

hub_class <- classify_hub_conservation(
  list(BDIS = hubs_bdis, BSYL = hubs_bsyl), hub_trait,
  module_comparisons = list("BDIS.BSYL" = mod_comp)
)

show_table(hub_class |> filter(classification != "non_hub"),
           caption = "Hub conservation (BDIS vs BSYL)")
```

### Co-expression partners of hub HOGs

After identifying trait-specific hubs, the natural question is: "Which
other HOGs co-express with this hub across species?"
`get_coexpressed_hogs()` resolves a HOG to its gene(s) in each species,
extracts co-expression neighbors from the network, maps them back to
HOGs, and aggregates across species. The `coexpressed_traits` column
makes trait-group filtering straightforward:

```{r hub-partners}
# Pick a trait-specific hub (if any exist)
ts_hubs <- hub_class |>
  filter(classification %in% c("trait_specific_hub", "rewired_hub",
                                "conserved_hub"))

if (nrow(ts_hubs) > 0) {
  candidate <- ts_hubs |> pull(hog) |> first()
  cat("Querying partners of", candidate, "\n\n")

  partners <- get_coexpressed_hogs(
    candidate, networks, orthologs,
    species_trait = trait_vec, min_species = 2L,
    edges = edges
  )

  cat(nrow(partners), "partner HOGs co-expressed in >= 2 species\n")

  if (nrow(partners) > 0) {
    annual_only <- partners |>
      filter(!is.na(coexpressed_traits), coexpressed_traits == "annual")
    cat(nrow(annual_only), "annual-only partners\n")

    cross_trait <- partners |>
      filter(!is.na(coexpressed_traits),
             stringr::str_detect(coexpressed_traits, ","))
    cat(nrow(cross_trait), "cross-trait partners\n")

    partners |>
      slice_head(n = 20) |>
      show_table(caption = paste("Top co-expression partners of", candidate))
  }
} else {
  cat("No trait-specific hubs found in BDIS vs BSYL comparison\n")
}
```

## 5. Edge robustness: density sweep

**Network density** controls how many edges are retained in the
co-expression network. At 3% density, only the top 3% of all
pairwise MR values pass the threshold -- these are the strongest
co-expression relationships.

We sweep over 11 density levels centered at the 3% baseline,
testing both directions:

- **Tighter** (below 3%): does the signal survive stricter
  thresholds? Edges that persist are strong, not marginal.
- **Looser** (above 3%): is the signal specific to this density,
  or stable across a range? Cliques that only appear at very loose
  thresholds are likely noise from extra edges.

The 3% choice is somewhat arbitrary. Centering the sweep around it
reveals sensitivity to that choice -- a truly robust co-expressolog
should be present across a range of densities, not just at one
particular cutoff.

The permutation pipeline is re-run at each level because network
neighborhoods change with the threshold, altering the HOG-level
q-values:

```{r density-sweep}
sweep <- tibble(multiplier = seq(0.95, 1.05, by = 0.01)) |>
  mutate(results = map(multiplier, \(m) {
    tight_nets <- map(networks, \(net)
      list(network = net$network, threshold = net$threshold * m))

    eff_density <- mean(map_dbl(networks, \(net) {
      vals <- net$network[upper.tri(net$network)]
      mean(vals >= net$threshold * m)
    }))

    tight_edges <- pairs |>
      mutate(edges = map2(sp1, sp2, \(a, b) {
        comp <- tryCatch(
          compare_neighborhoods(tight_nets[[a]], tight_nets[[b]], orthologs),
          error = \(e) NULL)
        if (is.null(comp) || nrow(comp) == 0) return(NULL)
        hog_res <- permutation_hog_test(
          tight_nets[[a]], tight_nets[[b]], comp,
          n_cores = n_cores, use_torch = TRUE)
        build_perm_edges(comp, hog_res, a, b)
      })) |>
      pull(edges) |>
      compact() |>
      list_rbind()

    if (is.null(tight_edges) || nrow(tight_edges) == 0) {
      tight_edges <- edges[0, ]
    }

    list(edges = tight_edges, eff_density = eff_density,
         n_conserved = sum(tight_edges$type == "conserved"))
  })) |>
  unnest_wider(results)

# Edge survival across density levels
sweep |>
  transmute(
    Multiplier = paste0(multiplier, "x"),
    Eff_density = sprintf("%.2f%%", eff_density * 100),
    Conserved_edges = n_conserved,
    Baseline = if_else(multiplier == 1, "***", "")
  ) |>
  show_table(caption = "Conserved edges across density levels")
```

The edge count may not decrease monotonically because the permutation
test operates at the HOG level: tighter neighborhoods can sharpen the
signal for some HOGs while eliminating it for others. Local fluctuations
reflect the discrete nature of the permutation q-values. The
interpretation is asymmetric: surviving tighter thresholds indicates
strong signal, while surviving looser thresholds indicates specificity
(not merely noise from additional edges).

**Analytical alternative: `clique_persistence()`.**
The density sweep above is rigorous but expensive — it re-runs the
full permutation pipeline at each threshold level. For a quick estimate,
`clique_persistence()` computes the ratio of each co-expressolog edge's
MR value to the species' density threshold. A persistence of 1.5 means
the weakest supporting edge is 50% above threshold and would survive a
1.5x tightening. This is a principled analytical proxy: it correctly
identifies the weakest link in each clique's co-expressolog support.
However, it cannot detect that tighter neighborhoods may change which
HOGs reach significance in the permutation test, or that new cliques
may form from HOGs that become significant at stricter thresholds.
Use persistence for screening or when GPU is unavailable; use the full
sweep for publication-quality robustness assessment.

## 6. Clique detection and structural stability

Using edges from the baseline density level, we find cliques and
test their structural stability via leave-k-out jackknife. The
jackknife removes k species at a time and checks whether each
clique's gene assignments are preserved (purely structural,
trait-agnostic). Trait annotations are added post-hoc:

```{r cliques-stability}
# Use the baseline (1.0x) edges for clique detection and stability
robust_edges <- sweep |> filter(multiplier == 1) |> pull(edges) |> pluck(1)
cat("Using", sum(robust_edges$type == "conserved"), "conserved edges at",
    sprintf("%.2f%%",
            (sweep |> filter(multiplier == 1) |> pull(eff_density)) * 100),
    "density\n\n")

# Find cliques
cliques <- find_cliques(robust_edges, species, min_species = 2L)
cat(nrow(cliques), "cliques\n")

cliques |>
  count(n_species) |>
  show_table(pageLength = 10, caption = "Cliques by number of species")
```

```{r stability}
# Jackknife stability (all k from 1 to N-2)
max_k <- length(species) - 2L
stab <- clique_stability(robust_edges, species,
                          species_trait = trait_vec,
                          all_species = species,
                          full_cliques = cliques,
                          max_k = max_k)
cat("Cliques tested:", length(stab$stability_class), "\n")
cat("Structurally stable at k=1:", sum(stab$stability_class >= 1L), "\n")

tibble(stability_class = stab$stability_class) |>
  mutate(stability_class = factor(stability_class, levels = 0:max_k)) |>
  count(stability_class, .drop = FALSE) |>
  show_table(pageLength = 10, caption = "Stability class distribution")
```

## 7. HOG classification with robustness ranking

Classify each HOG by its conservation pattern, then rank within each
category by robustness. The waterfall assigns each HOG to exactly one
class: complete > partial > differentiated > trait_specific > unclassified.

```{r classify}
result <- classify_cliques(robust_edges, species, trait_vec,
                            min_species = 2L,
                            stability = stab)

result |>
  count(classification) |>
  show_table(pageLength = 5, caption = "HOG classification summary")
```

- **complete**: all 8 species form a clique -- deeply conserved across
  all Pooideae
- **partial**: clique spans both annual and perennial species, but not
  all 8
- **differentiated**: annual species have a clique AND perennial species
  have a clique, but no cross-group conservation -- lineage-specific
  co-expression programs
- **trait_specific**: only annuals or only perennials have a clique
- **unclassified**: no conservation signal detected

The robustness ranking comes from combining density sweep survival
with stability class. HOGs classified at a stricter density level
with high stability class are the strongest findings:

```{r ranking}
# Count how many density levels each HOG's edges survive
hog_survival <- sweep |>
  transmute(multiplier,
            hog = map(edges, \(e)
              e |> filter(type == "conserved") |> distinct(hog) |> pull(hog))) |>
  unnest(hog) |>
  count(hog, name = "n_densities_survived")

result <- result |>
  left_join(hog_survival, by = "hog") |>
  mutate(n_densities_survived = replace_na(n_densities_survived, 0L))

# Annotate trait composition from the best clique per HOG
hog_traits <- cliques |>
  group_by(hog) |>
  slice_min(mean_q, n = 1, with_ties = FALSE) |>
  ungroup() |>
  pivot_longer(all_of(species), names_to = "sp", values_to = "gene") |>
  filter(!is.na(gene)) |>
  summarise(
    n_annual = sum(trait_vec[sp] == "annual"),
    n_perennial = sum(trait_vec[sp] == "perennial"),
    .by = hog
  )

result <- left_join(result, hog_traits, by = "hog")

# Show ranked HOGs per category
show_cols <- c("hog", "n_species", "n_annual", "n_perennial",
               "best_mean_q", "stability_class", "n_densities_survived")
show_cols <- intersect(show_cols, names(result))
```

Every clique now has a `stability_class` -- the highest k at which it
survives ALL species-removal subsets. This is a purely structural measure.
Trait annotations (from the `traits` column in the stability output) are
orthogonal: they describe which life-cycle groups each clique spans, not
whether the clique is robust.

### {.tabset}

#### Partial

```{r partial-table}
show_classification(result, "partial", show_cols,
                    "Partial HOGs (cross-trait cliques)")
```

#### Trait-specific

```{r trait-specific-table}
show_classification(result, "trait_specific", show_cols,
                    "Trait-specific HOGs (one lineage only)")
```

#### Differentiated

```{r differentiated-table}
show_classification(result, "differentiated", show_cols,
                    "Differentiated HOGs (within-group conservation, cross-group divergence)")
```

#### Complete

```{r complete-table}
show_classification(result, "complete", show_cols,
                    "Complete HOGs (all 8 species)")
```

###

## Session info

```{r sessioninfo}
sessionInfo()
```
