---
title: "Walkthrough: the building blocks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Walkthrough: the building blocks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_se <- requireNamespace("SummarizedExperiment", quietly = TRUE)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = has_se)
tab <- function(df, ...) knitr::kable(df, digits = 3, row.names = FALSE, ...)
```

The [quickstart](quickstart.html) runs `rcomplex()` in one call. This
article runs the same analysis step by step on eight Pooideae grasses.
Use it when you need to change one step, to inspect an intermediate
object, or to run the module and trait tests.

| Species | Code | Life cycle | Genus pair |
|---------|------|------------|------------|
| *Brachypodium distachyon* | BDIS | annual | Brachypodium |
| *Brachypodium sylvaticum* | BSYL | perennial | Brachypodium |
| *Hordeum vulgare* | HVUL | annual | Hordeum |
| *Hordeum jubatum* | HJUB | perennial | Hordeum |
| *Briza maxima* | BMAX | annual | Briza |
| *Briza media* (mapped to BMAX) | BMED | perennial | Briza |
| *Vulpia bromoides* | VBRO | annual | Festuca/Vulpia |
| *Festuca pratensis* | FPRA | perennial | Festuca/Vulpia |

The data are VST-normalised RNA-seq for the 1,500 most expressed hogs
from the "Live or let die?" project, stored as
`SummarizedExperiment` objects. Each species has 20 leaf and 20 root
samples.

## 1. Inputs

We keep the 20 leaf samples. A network built from pooled leaf and root
samples mostly separates tissues, so use one tissue per network or
remove the tissue with `block` (see the methods article).

```{r data, message = FALSE}
library(rcomplex)
library(SummarizedExperiment)

se_list <- readRDS(system.file("extdata", "pooideae_vignette.rds",
  package = "rcomplex"
))
se_list <- lapply(se_list, function(s) s[, s$tissue == "leaf"])
species <- names(se_list)

clades <- list(
  annual = c("BDIS", "HVUL", "BMAX", "VBRO"),
  perennial = c("BSYL", "HJUB", "BMED", "FPRA")
)
```

`clades` is a named list of species vectors. Every function that takes
a trait or a lineage takes this one shape. `clades_from_tree()` builds
it from a species tree.

The ortholog table has one row per gene: `species`, `gene` and `hog`.
Here it comes from `rowData()`. For an OrthoFinder or PLAZA file, call
`read_orthologs()`.

```{r hogs}
long <- do.call(rbind, lapply(species, function(s) {
  data.frame(
    species = s, gene = rownames(se_list[[s]]),
    hog = as.character(rowData(se_list[[s]])$hog)
  )
}))
long <- long[!is.na(long$hog), ]
tab(head(long, 4))
```

### Paralog reduction

Recent duplicates often have near-identical expression. They inflate
the number of cliques without adding information.
`reduce_orthogroups()` merges the paralogs of a hog that correlate
above `cor_threshold` and keeps divergent copies apart.

```{r reduce}
reduced <- lapply(species, function(s) {
  reduce_orthogroups(
    assay(se_list[[s]], 1), long[long$species == s, c("gene", "hog")],
    gene_col = "gene", cor_threshold = 0.7
  )
})
names(reduced) <- species
vapply(reduced, function(r) r$n_merged, integer(1))
```

`prepare_orthologs()` turns the long table into ortholog pairs and maps
the genes through the reduction.

```{r orthologs}
orthologs <- prepare_orthologs(long, reduced)
nrow(orthologs)
```

## 2. Networks

`compute_network()` correlates all gene pairs and converts the
correlations to mutual rank (MR). MR asks whether two genes rank each
other among their best partners, so a gene that correlates with
everything does not dominate. The network keeps the top 3% of gene
pairs as edges.

The sparse store holds more pairs than the analysis threshold:
`store_density = 0.06` leaves room for the density sweep in section 4.

```{r networks}
networks <- lapply(reduced, function(r) {
  compute_network(r$expr_matrix, density = 0.03, store_density = 0.06)
})
```

Each network records the correlation it needed to pass the threshold.
At 20 samples this value is low, and many edges are chance
correlations.

```{r rthreshold}
vapply(networks, function(n) n$params$r_threshold, numeric(1))
```

Networks built elsewhere enter through `as_network()`. Pass a gene by
gene matrix of edge weights or a sparse matrix.

## 3. Co-expressologs

`find_coexpressologs()` tests every ortholog pair in every species pair.
It maps the neighbours of gene A in species 1 to species 2 and asks how
many are neighbours of the ortholog of A. The test runs in both
directions, and both must pass.

```{r edges}
edges <- find_coexpressologs(networks, orthologs, seed = 1)
table(edges$type)
tab(head(edges[order(-edges$effect_size), c(
  "gene1", "gene2", "hog", "score", "evalue", "q_value", "effect_size",
  "power"
)], 4))
```

- `score` is `-log2(p)` and `evalue` is `n_tests * p`.
- `effect_size` is the fold enrichment of shared neighbours.
- `power` is the chance to call the pair had it been conserved. A
  non-significant pair with low power is weak evidence of divergence.

`type == "conserved"` marks a co-expressolog.

### Hypergeometric or rank

The default test is the hypergeometric test of ComPlEx. It is fast and
matches the published method. `method = "rank"` ranks every gene in
species 2 by its co-expression with the translated neighbours, and
reports the ortholog's place in that ranking. The rank test needs the
argument `null_networks`: one network per species, built by
`null_network()` from the expression with the samples shuffled.

### Nulls

`coexpressolog_null()` rewires each network with the degrees held fixed
and repeats the test. It answers how many calls to expect from the
degree sequence alone.

```{r null, eval = FALSE}
nul <- coexpressolog_null(networks, orthologs, n_cores = 2, seed = 1)
```

`rcomplex(null = TRUE)` runs the cheaper shuffled-expression null and
reports `calls_null` per species pair.

## 4. Density sweep

A 3% density is a convention. A conserved pair should survive a range
of densities. `density_sweep()` repeats the test at a set of
multipliers of the threshold. A multiplier above 1 tightens the network
and a multiplier below 1 loosens it. The loosest multiplier must stay
inside the stored edges.

```{r sweep, message = FALSE}
multipliers <- seq(0.975, 1.025, by = 0.0125)
store_floor <- max(vapply(
  networks, function(n) n$store_threshold / n$threshold, numeric(1)
))
stopifnot(min(multipliers) >= store_floor)

sweep <- density_sweep(networks, orthologs,
  multipliers = multipliers, method = "hypergeometric", seed = 1
)
tab(sweep[, c("multiplier", "eff_density", "n_significant")])
```

Raw MR values are compressed near the top. A small multiplier moves the
effective density a lot. Count, for each hog, how many levels keep a
conserved edge:

```{r survival}
survive <- table(unlist(lapply(sweep$edges, function(e) {
  unique(e$hog[e$type == "conserved"])
})))
table(factor(survive, levels = seq_len(nrow(sweep))))
```

## 5. Modules

A module is a group of genes that are co-expressed with each other.
`detect_modules()` runs Leiden clustering. Check that your modules
replicate before you test them. At 20 samples they often do not.

```{r modules}
modules <- lapply(networks, detect_modules, seed = 42)
vapply(modules, function(m) m$n_modules, integer(1))
```

`as_modules()` builds the same object from any gene partition, such as
the output of another tool.

### Preservation

`preservation_paired()` asks, for each ordered species pair, whether
the modules of the reference species keep their wiring in the partner
network. It tests two statistics against a gene-identity permutation
null: the density of the module (`avg.weight`) and the correlation of
intramodular connectivity (`cor.degree`). A module is preserved only
when both are significant. Each module is `conserved`, `moderate`,
`diverged` or `untested`.

The trait test in the next section needs all pairs. Pairs between genera
supply the same-trait contrasts. `n_perm = 500` keeps the example
short. Use the default of 10,000 for results you report.

```{r preservation, message = FALSE}
pairs <- as.data.frame(t(utils::combn(species, 2)))
names(pairs) <- c("species1", "species2")

trait <- stack(clades)
trait <- setNames(as.character(trait$ind), trait$values)

pres <- preservation_paired(modules, networks, orthologs,
  pairs = pairs, group = trait, edges = edges, n_perm = 500, seed = 42
)
table(pres$classification$classification)
```

### Is divergence associated with the trait?

`preservation_matrix_test()` asks one question of the whole matrix:
are trait-discordant pairs the less preserved ones? It compares the
mean `Zsummary_std` of concordant pairs (two annuals or two
perennials) with that of discordant pairs. It refers the difference to
a null that relabels the traits of the species.

`block` names the genera. It enables a second null that relabels only
within a genus, which holds the phylogeny fixed. It also drops the
within-genus pairs, where trait and relatedness are confounded.

```{r matrix-test}
genus <- list(
  Brachypodium = c("BDIS", "BSYL"), Hordeum = c("HVUL", "HJUB"),
  Briza = c("BMAX", "BMED"), Festuca = c("VBRO", "FPRA")
)
pmt <- suppressWarnings(preservation_matrix_test(
  pres$classification, trait,
  block = genus, seed = 1
))
tab(pmt$class_means)
c(
  observed = pmt$observed, p_free = pmt$p_free, p_blocked = pmt$p_blocked,
  floor_free = pmt$free$p_attainable,
  floor_blocked = pmt$blocked$p_attainable
)
```

A positive statistic means discordant pairs are less preserved. On this
data the statistic is near zero, so the leaf modules carry no trait
signal. The two nulls agree.

The floor of a relabelling p-value comes from the species set. Eight
species split 4/4 give `choose(8, 4) = 70` labellings, and two of them
share the maximum, so the floor is `2/70 = 0.029`. Within four genera
the floor is `2/16 = 0.125`. No amount of sampling lowers either floor.
`preservation_matrix_test()` warns when the floor exceeds 0.05, and the
chunk above silences that warning.

### P-value resolution

A permutation p-value cannot fall below `1 / (n_perm + 1)`. Tests whose
true p-value is smaller all return the floor and tie. Rank on
`Zsummary_std` and keep p and q for the significance call.
`pvalue_resolution()` reports the ties:

```{r resolution}
pvalue_resolution(pres$classification$q_value, n_perm = 500)
```

## 6. Hubs

`identify_module_hubs()` finds the hub genes of each module. The
primary centrality is the weighted degree. It also reports betweenness
and eigenvector centrality.

```{r hubs}
hubs <- lapply(species, function(s) {
  identify_module_hubs(modules[[s]], networks[[s]], orthologs)
})
names(hubs) <- species
vapply(hubs, function(h) sum(h$is_hub), integer(1))
```

`classify_hub_conservation()` labels each hog by where it is a hub.
It needs `module_correspondence()`, which tests which module of species
1 matches which module of species 2. Reuse the ortholog map that
`preservation_paired()` resolved. The key is the sorted species pair.

```{r hub-conservation}
pairs_genus <- lapply(genus, sort)
correspondence <- lapply(pairs_genus, function(sp) {
  key <- paste(sp, collapse = ".")
  module_correspondence(modules[[sp[1]]], modules[[sp[2]]],
    pres$raw[[key]]$map,
    species_ref = sp[1], species_test = sp[2], seed = 1
  )
})
names(correspondence) <- vapply(
  pairs_genus, paste, character(1), collapse = "."
)

hub_class <- classify_hub_conservation(hubs, clades,
  module_comparisons = correspondence
)
table(hub_class$classification)
```

`get_coexpressed_hogs()` lists the hogs that co-express with a
candidate hub in several species.

```{r hub-partners}
candidate <- hub_class$hog[hub_class$classification == "conserved_hub"][1]
partners <- get_coexpressed_hogs(candidate, networks, orthologs,
  clades = clades, min_species = 2L, edges = edges
)
tab(head(partners[, c(
  "partner_hog", "n_species", "coexpressed_traits", "partner_min_q"
)], 5))
```

## 7. Cliques

A clique is a set of genes, one per species, that are all co-expressed
with each other. The edges of one hog form a graph, and rcomplex finds
cliques on it in two ways. The two answer different questions.

### Species graph

`find_cliques()` cliques the graph of species. It returns one best gene
assignment per clique. Give it the full edge table, so the intensity
scale uses every tested pair.

```{r cliques}
cliques <- find_cliques(edges, species, min_species = 2L)
table(cliques$n_species)
```

A clique of two species is a single conserved edge. Cliques that span
three or more species need conservation across independent lineages.

`clique_stability()` removes k species at a time and checks whether each
clique persists. A clique that survives the removal of any one species
does not depend on a single lineage.

```{r stability}
stab <- clique_stability(edges, species,
  clades = clades, all_species = species,
  full_cliques = cliques, max_k = 2L
)
table(stab$stability_class)
```

`classify_cliques()` assigns each hog to one class: `complete` (all
species), `partial` (both clades, not all species), `differentiated`
(a clique in each clade and none across), `trait_specific` (one clade
only) or `unclassified`. The `underpowered` column flags a
`differentiated` or `trait_specific` call that rests on edges with low
power.

```{r classify}
result <- classify_cliques(edges, species, clades,
  min_species = 2L, stability = stab
)
table(result$classification)
```

### Gene graph

`gene_clique_graph()` cliques the graph of genes. A hog with paralogs
gives one clique per gene combination. `classify_gene_cliques()` applies
the six-tier taxonomy of Rodriguez *et al.* (2026): `complete_conserved`,
`lineage_specific`, `partial_significant`, `partial_present`,
`differentiated` and `trait_specific`. A clique that fits none is
`unclassified`. A tier that rests on low power becomes `underpowered`.

```{r gene-cliques}
gene_cliques <- gene_clique_graph(edges)
gene_class <- classify_gene_cliques(gene_cliques, edges, species,
  clades = clades
)
table(gene_class$classification)
```

Pass the unfiltered edge table. The gap tiers need the pairs that were
tested and failed, so they can refuse them.

Use the species graph for stability and for one clique per hog. Use the
gene graph to see every paralog combination and to follow the published
taxonomy. The species graph finds more `trait_specific` hogs because it
asks for a clique among the species it can see, and the gene graph asks
for a complete clique of one clade with an outside species tested at
adequate power.

## 8. Reading the results

Combine the robustness signals. A hog with a `complete` class, a
stability class of 2 or more, and survival at most density levels is a
strong candidate.

```{r robust}
result$n_densities <- as.integer(survive[result$hog])
result$n_densities[is.na(result$n_densities)] <- 0L
strong <- result[result$classification == "complete" &
  result$stability_class >= 2L &
  result$n_densities >= nrow(sweep) - 1L, ]
nrow(strong)
tab(head(strong[order(strong$best_mean_q), c(
  "hog", "n_species", "best_mean_q", "stability_class", "n_densities"
)], 5))
```

Check three things before you believe any result.

1. Read the correlation each network needed (`r_threshold`).
2. Read the power of the edges behind a negative call.
3. Read `pvalue_resolution()` for every permutation p-value.

The [methods article](methods.html) gives the statistics behind each
step.
