---
title: "Quickstart"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quickstart}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_se <- requireNamespace("SummarizedExperiment", quietly = TRUE)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = has_se)
```

This quickstart runs the whole analysis with one call, `rcomplex()`.
It answers one question: which genes keep their co-expression partners
in other species?

## Data

The package ships leaf and root RNA-seq from eight Pooideae grasses.
Each species has 20 leaf samples. We use four species: two annual and
two perennial.

```{r data, message = FALSE}
library(rcomplex)
library(SummarizedExperiment)

se <- readRDS(system.file("extdata", "pooideae_vignette.rds",
  package = "rcomplex"
))
leaf <- lapply(
  se[c("BDIS", "BSYL", "HVUL", "HJUB")],
  function(s) s[, s$tissue == "leaf"]
)
```

`rcomplex()` needs two inputs. The first is one expression object per
species, with genes in rows and samples in columns. The second is a
table of hogs: ortholog groups with the columns `species`, `gene` and
`hog`. For an OrthoFinder file, call `read_orthologs()` instead.

```{r hogs}
hogs <- do.call(rbind, lapply(names(leaf), function(s) {
  data.frame(
    species = s, gene = rownames(leaf[[s]]),
    hog = rowData(leaf[[s]])$hog
  )
}))
head(hogs, 3)
```

## Run

`clades` names the species sets. `null = TRUE` repeats the comparison
on shuffled networks. The run takes about 10 seconds.

```{r run}
res <- rcomplex(leaf, hogs,
  clades = list(
    annual = c("BDIS", "HVUL"),
    perennial = c("BSYL", "HJUB")
  ),
  null = TRUE, seed = 1
)
res
```

The header lists, per species, the genes, the samples and the density.
The value after "r >=" is the weakest correlation that became an edge. At 20 samples,
a correlation of 0.3 is weak evidence. Read this value first.

The last line counts the cliques by tier. A clique is a set of genes,
one per species, that are all co-expressed. A hog with paralogs gives one clique per gene combination.

## Read the results

`summary()` gives the tier table. With `null = TRUE` it also gives the
calls on shuffled networks.

```{r summary}
summary(res)
```

`calls_null` counts the calls that survive on shuffled networks. Zero
means the false-call rate at this `alpha` is low on this data.

`res$edges` has one row per ortholog pair and species pair.

```{r edges}
head(res$edges[, c(
  "gene1", "gene2", "hog", "score", "evalue",
  "q_value", "effect_size", "power"
)], 4)
```

- `score` is `-log2(p)`. Higher is stronger.
- `evalue` is the expected number of pairs this strong by chance.
- `q_value` is the false discovery rate.
- `effect_size` is the fold enrichment of shared partners.
- `power` is the chance to call the pair if it were conserved.

A low `power` means a non-significant pair is weak evidence of
divergence.

`res$classification` has one row per clique. `as.data.frame(res)`
returns the same table.

```{r tiers}
table(as.data.frame(res)$classification)
```

## Save

`write_rcomplex()` writes `edges.tsv`, `cliques.tsv` and
`classification.tsv`. After a null run it adds `edges_null.tsv`.

```{r save}
out <- file.path(tempdir(), "rcomplex-out")
write_rcomplex(res, out)
list.files(out)
```

## Check the p-value floor

Run `pvalue_resolution()` on any p-values you report. It shows how many
values tie at the floor.

```{r resolution}
pvalue_resolution(res$edges$p_value)
```

## Next steps

- Designed experiment: pass `block` to remove a designed factor.
- Networks built elsewhere: pass `networks = ` to `rcomplex()`.
- Modules: pass `modules = TRUE`.
- Each step alone: see the walkthrough article.
