---
title: "Candidate heterochrony and network context"
subtitle: "Clique-resolved annual-perennial rewiring in Pooideae"
date: last-modified
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    page-layout: full
    embed-resources: true
    smooth-scroll: true
execute:
  echo: false
  warning: false
  message: false
---

<style>
:root {
  --annual: #d95f02;
  --perennial: #167d8d;
  --ink: #27323a;
  --muted: #65737e;
}
.quarto-title-banner {
  background: linear-gradient(120deg, #16324f, #167d8d);
  color: white;
}
.metric-strip {
  display: grid;
  grid-template-columns: repeat(auto-fit, minmax(145px, 1fr));
  gap: 0.8rem;
  margin: 1rem 0 1.5rem;
}
.metric-card {
  border: 1px solid #dde4e8;
  border-radius: 0.6rem;
  padding: 0.85rem 1rem;
  background: #f8fafb;
  box-shadow: 0 1px 3px rgb(0 0 0 / 6%);
}
.metric-value {
  color: var(--ink);
  font-size: 1.55rem;
  font-weight: 700;
}
.metric-label {
  color: var(--muted);
  font-size: 0.82rem;
  line-height: 1.2;
}
.annual-text { color: var(--annual); font-weight: 650; }
.perennial-text { color: var(--perennial); font-weight: 650; }
.gt_table { font-size: 0.88rem; }
</style>

```{r}
#| label: setup

library(ggplot2)

find_root <- function() {
  candidates <- c(".", "..")
  found <- candidates[file.exists(file.path(candidates, "DESCRIPTION"))]
  if (length(found) == 0L) {
    stop("Could not locate the rcomplex repository root")
  }
  normalizePath(found[1L])
}

root <- find_root()
output_path <- function(name) {
  file.path(root, "analysis", "output", name)
}
cache_path <- function(name) {
  file.path(root, "analysis", "cache", "rewiring-screen", name)
}

safe_rbind <- function(x) {
  x <- x[lengths(x) > 0L]
  if (length(x) == 0L) {
    return(data.frame())
  }
  result <- do.call(rbind, x)
  rownames(result) <- NULL
  result
}

candidates <- read.csv(
  output_path("rewiring-heterochrony-candidates.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
profiles <- read.csv(
  output_path("rewiring-heterochrony-candidate-profiles.csv"),
  stringsAsFactors = FALSE
)
literature <- read.csv(
  output_path("rewiring-hot-hog-literature.csv"),
  stringsAsFactors = FALSE
)
selected <- read.csv(
  cache_path("selected-anchors.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
regulator_copies <- read.csv(
  output_path("rewiring-heterochrony-regulator-copies.csv"),
  stringsAsFactors = FALSE
)
membership <- read.csv(
  cache_path("module-membership.csv"),
  stringsAsFactors = FALSE
)

annual_colour <- "#d95f02"
perennial_colour <- "#167d8d"
neutral_colour <- "#7a8790"

first_nonempty <- function(...) {
  values <- list(...)
  result <- rep(NA_character_, max(lengths(values)))
  for (value in values) {
    value <- rep(value, length.out = length(result))
    use <- is.na(result) & !is.na(value) & nzchar(value)
    result[use] <- value[use]
  }
  result
}

short_text <- function(x, width = 52L) {
  x <- ifelse(is.na(x), "", x)
  ifelse(
    nchar(x) > width,
    paste0(substr(x, 1L, width - 1L), "\u2026"),
    x
  )
}

candidate_annotation <- first_nonempty(
  candidates$best_arabi_defline,
  candidates$regulator_arabi_defline,
  candidates$best_rice_defline,
  candidates$regulator_rice_defline
)
candidates$annotation <- candidate_annotation

anchor_genes <- selected[c("hog", "BDIS")]
names(anchor_genes) <- c("anchor_hog", "bdis_selected_gene")
extra_genes <- regulator_copies[
  regulator_copies$species == "BDIS" & regulator_copies$selected,
  c("anchor_hog", "selected_gene"),
  drop = FALSE
]
names(extra_genes)[2L] <- "bdis_selected_gene"
anchor_genes <- rbind(anchor_genes, extra_genes)
anchor_genes <- anchor_genes[!duplicated(anchor_genes$anchor_hog), ]
candidates <- merge(
  candidates,
  anchor_genes,
  by = "anchor_hog",
  all.x = TRUE,
  sort = FALSE
)

database_links <- function(gene) {
  ifelse(
    is.na(gene) | !nzchar(gene),
    "",
    paste0(
      "[Ensembl](",
      "https://plants.ensembl.org/Multi/Search/Results?q=",
      utils::URLencode(gene, reserved = TRUE),
      ") \u00b7 [UniProt](",
      "https://www.uniprot.org/uniprotkb?query=",
      utils::URLencode(gene, reserved = TRUE),
      ") \u00b7 [OrthoDB](",
      "https://www.orthodb.org/?query=",
      utils::URLencode(gene, reserved = TRUE),
      ")"
    )
  )
}

candidates$database_links <- database_links(
  candidates$bdis_selected_gene
)
candidates$display_label <- paste0(
  candidates$anchor_hog,
  ifelse(
    nzchar(candidates$annotation),
    paste0(" \u00b7 ", short_text(candidates$annotation, 35L)),
    ""
  )
)

pattern_labels <- c(
  coordinated_phase_shift = "Coordinated anchor + module",
  anchor_only_phase_shift = "Anchor only",
  module_only_phase_shift = "Module only",
  anchor_module_decoupling = "Anchor/module decoupled",
  no_recurrent_phase_shift = "No recurrent shift"
)
candidates$pattern_label <- unname(
  pattern_labels[candidates$heterochrony_pattern]
)
candidates$phase_intervals <- 4 * candidates$anchor_phase_difference
candidates$phase_label <- ifelse(
  candidates$anchor_phase_difference > 0,
  "Annual later",
  ifelse(candidates$anchor_phase_difference < 0, "Annual earlier", "Tied")
)

strong <- candidates$heterochrony_pattern != "no_recurrent_phase_shift"
n_hogs <- length(unique(candidates$anchor_hog))
n_strong_rows <- sum(strong)
n_coordinated <- sum(
  candidates$heterochrony_pattern == "coordinated_phase_shift"
)
n_later <- sum(candidates$phase_label == "Annual later" & strong)
n_earlier <- sum(candidates$phase_label == "Annual earlier" & strong)
```

::: {.callout-important}
## Interpretation boundary

Positive phase values mean later deployment in annuals on the matched
five-point sampling axis. They do not prove delayed developmental stage:
the samples lack an independent meristem-stage clock. Candidate selection
also used these expression data, so this is a conditional exploratory screen,
not a confirmatory significance analysis.
:::

```{r}
#| output: asis

cat(
  '<div class="metric-strip">',
  sprintf(
    paste0(
      '<div class="metric-card"><div class="metric-value">%d</div>',
      '<div class="metric-label">candidate HOGs</div></div>'
    ),
    n_hogs
  ),
  sprintf(
    paste0(
      '<div class="metric-card"><div class="metric-value">%d</div>',
      '<div class="metric-label">recurrent candidate-tissue shifts</div>',
      '</div>'
    ),
    n_strong_rows
  ),
  sprintf(
    paste0(
      '<div class="metric-card"><div class="metric-value">%d</div>',
      '<div class="metric-label">coordinated anchor/module shifts</div>',
      '</div>'
    ),
    n_coordinated
  ),
  sprintf(
    paste0(
      '<div class="metric-card"><div class="metric-value annual-text">',
      '%d</div><div class="metric-label">later annual profiles</div></div>'
    ),
    n_later
  ),
  sprintf(
    paste0(
      '<div class="metric-card"><div class="metric-value perennial-text">',
      '%d</div><div class="metric-label">earlier annual profiles</div></div>'
    ),
    n_earlier
  ),
  "</div>",
  sep = "\n"
)
```

## Strongest recurrent shifts

The primary phase statistic is an expression-weighted centroid on a normalized
axis from 0 (T1) to 1 (T5). One sampling interval is 0.25 units. The table is
sortable and searchable.

```{r}

top_table <- candidates[strong, c(
  "anchor_hog", "tissue", "pattern_label", "phase_label",
  "phase_intervals", "anchor_pair_agreement",
  "anchor_bootstrap_support", "module_phase_difference", "annotation"
)]
top_table <- top_table[order(
  top_table$pattern_label != "Coordinated anchor + module",
  -abs(top_table$phase_intervals)
), ]
top_table$annotation <- short_text(top_table$annotation, 42L)
gt::gt(top_table) |>
  gt::cols_label(
    anchor_hog = "HOG",
    tissue = "Tissue",
    pattern_label = "Evidence pattern",
    phase_label = "Direction",
    phase_intervals = "\u0394 phase (intervals)",
    anchor_pair_agreement = "Anchor pairs",
    anchor_bootstrap_support = "Bootstrap support",
    module_phase_difference = "Module \u0394 phase",
    annotation = "Annotation"
  ) |>
  gt::fmt_number(
    columns = c("phase_intervals", "module_phase_difference"),
    decimals = 2
  ) |>
  gt::fmt_percent(columns = "anchor_bootstrap_support", decimals = 0) |>
  gt::data_color(
    columns = "phase_intervals",
    palette = c(perennial_colour, "#f7f7f7", annual_colour),
    domain = c(-1, 1)
  ) |>
  gt::tab_header(
    title = "Candidate heterochrony calls",
    subtitle = "Ranked by evidence pattern and absolute anchor phase shift"
  ) |>
  gt::tab_source_note(
    gt::md(
      paste(
        "Pair support is out of four sister pairs.",
        "Positive phase means later annual deployment."
      )
    )
  ) |>
  gt::opt_interactive(
    use_sorting = TRUE,
    use_search = TRUE,
    use_filters = TRUE,
    use_compact_mode = TRUE,
    use_page_size_select = TRUE,
    page_size_default = 15
  )
```

## Distribution across tissues

```{r}

summary_input <- candidates[c(
  "tissue", "pattern_label", "phase_intervals",
  "anchor_pair_agreement", "anchor_bootstrap_support",
  "module_phase_difference"
)]
names(summary_input) <- c(
  "Tissue", "Evidence pattern", "Anchor phase shift",
  "Pair agreement", "Bootstrap support", "Module phase shift"
)

gtsummary::tbl_summary(
  summary_input,
  by = "Tissue",
  statistic = list(
    gtsummary::all_continuous() ~ "{median} ({p25}, {p75})",
    gtsummary::all_categorical() ~ "{n} ({p}%)"
  ),
  digits = gtsummary::all_continuous() ~ 2,
  missing = "no"
) |>
  gtsummary::add_overall(last = TRUE) |>
  gtsummary::bold_labels() |>
  gtsummary::modify_caption(
    "**Candidate-level timing summaries by tissue**"
  ) |>
  gtsummary::as_gt()
```

## Expression-profile atlas

Thin lines are individual species; thick lines are life-history means. Values
are robustly standardized within each species and therefore describe
deployment shape rather than absolute cross-species transcript abundance.
Solid lines represent the clique-resolved candidate gene and dashed lines its
recurrent local module.

```{r}
#| fig-width: 15
#| fig-height: 12
#| fig-alt: >
#|   Faceted expression trajectories for candidate genes and their modules,
#|   comparing annual and perennial species over five matched sampling points.

profile_meta <- unique(candidates[c(
  "anchor_hog", "display_label", "heterochrony_pattern"
)])
profile_meta <- profile_meta[
  !duplicated(profile_meta$anchor_hog),
  ,
  drop = FALSE
]
profile_data <- merge(
  profiles,
  profile_meta,
  by = "anchor_hog",
  all.x = TRUE,
  sort = FALSE
)
life_cycle <- c(
  BDIS = "Annual", BSYL = "Perennial",
  HVUL = "Annual", HJUB = "Perennial",
  BMAX = "Annual", BMED = "Perennial",
  VBRO = "Annual", FPRA = "Perennial"
)
profile_data$life_cycle <- unname(life_cycle[profile_data$species])
profile_data$layer <- factor(
  profile_data$layer,
  levels = c("anchor", "module"),
  labels = c("Candidate", "Module")
)
profile_data$life_cycle <- factor(
  profile_data$life_cycle,
  levels = c("Annual", "Perennial")
)

profile_order <- aggregate(
  abs(anchor_phase_difference) ~ anchor_hog,
  candidates,
  max
)
profile_order <- profile_order[order(
  -profile_order$`abs(anchor_phase_difference)`
), ]
profile_hogs <- profile_order$anchor_hog
profile_pages <- split(
  profile_hogs,
  ceiling(seq_along(profile_hogs) / 7L)
)

plot_profile_page <- function(hogs) {
  plot_data <- profile_data[
    profile_data$anchor_hog %in% hogs,
    ,
    drop = FALSE
  ]
  plot_data$display_label <- factor(
    plot_data$display_label,
    levels = profile_meta$display_label[
      match(hogs, profile_meta$anchor_hog)
    ]
  )
  means <- aggregate(
    value ~ display_label + tissue + stage_position + life_cycle + layer,
    plot_data,
    mean
  )
  ggplot(
    plot_data,
    aes(
      x = stage_position,
      y = value,
      colour = life_cycle,
      linetype = layer
    )
  ) +
    geom_line(
      aes(group = interaction(species, layer)),
      alpha = 0.22,
      linewidth = 0.45
    ) +
    geom_line(
      data = means,
      aes(group = interaction(life_cycle, layer)),
      linewidth = 1.05
    ) +
    geom_point(data = means, size = 1.2) +
    facet_grid(tissue ~ display_label) +
    scale_colour_manual(
      values = c(Annual = annual_colour, Perennial = perennial_colour)
    ) +
    scale_linetype_manual(values = c(Candidate = "solid", Module = "22")) +
    scale_x_continuous(
      breaks = seq(0, 1, by = 0.25),
      labels = paste0("T", 1:5)
    ) +
    labs(
      x = "Matched sampling position",
      y = "Within-species robust z-score",
      colour = NULL,
      linetype = NULL
    ) +
    theme_minimal(base_size = 10) +
    theme(
      panel.grid.minor = element_blank(),
      strip.text.x = element_text(size = 7.5),
      legend.position = "top",
      axis.text.x = element_text(size = 7)
    )
}

for (page in profile_pages) {
  print(plot_profile_page(page))
}
```

## Annual and perennial ego-networks

These views hold each candidate anchor fixed and count in how many annual or
perennial species each recurrent HOG is an immediate network neighbour. They
show neighbour recruitment, not regulation direction and not edges among
neighbours.

```{r}
#| fig-width: 15
#| fig-height: 8
#| fig-alt: >
#|   Annual and perennial candidate-centred ego networks, with edge width
#|   indicating recurrence across four species and node colour indicating
#|   which life-history group has stronger support.

coordinated <- candidates[
  candidates$heterochrony_pattern == "coordinated_phase_shift" &
    candidates$is_eligible_anchor,
  ,
  drop = FALSE
]
ego_rank <- aggregate(
  abs(anchor_phase_difference) ~ anchor_hog,
  coordinated,
  max
)
ego_rank <- ego_rank[order(-ego_rank$`abs(anchor_phase_difference)`), ]
ego_hogs <- c(
  "HOG:0025149",
  utils::head(setdiff(ego_rank$anchor_hog, "HOG:0025149"), 7L)
)

membership$life_cycle <- unname(life_cycle[membership$species])
ego_membership <- unique(membership[
  membership$anchor_hog %in% ego_hogs &
    membership$module_hog != membership$anchor_hog,
  c("anchor_hog", "module_hog", "species", "life_cycle", "status")
])
ego_membership$is_neighbor <- ego_membership$status == "anchor_neighbor"
ego_support <- aggregate(
  is_neighbor ~ anchor_hog + module_hog + life_cycle,
  ego_membership,
  sum
)
names(ego_support)[names(ego_support) == "is_neighbor"] <- "support"
ego_support <- reshape(
  ego_support,
  idvar = c("anchor_hog", "module_hog"),
  timevar = "life_cycle",
  direction = "wide"
)
for (column in c("support.Annual", "support.Perennial")) {
  if (!column %in% names(ego_support)) {
    ego_support[[column]] <- 0
  }
  ego_support[[column]][is.na(ego_support[[column]])] <- 0
}
ego_support$difference <- ego_support$support.Annual -
  ego_support$support.Perennial
ego_support$bias <- ifelse(
  ego_support$difference >= 2,
  "Annual-enriched",
  ifelse(
    ego_support$difference <= -2,
    "Perennial-enriched",
    "Shared or weak"
  )
)
ego_support <- ego_support[
  pmax(ego_support$support.Annual, ego_support$support.Perennial) >= 3 |
    abs(ego_support$difference) >= 2,
  ,
  drop = FALSE
]
ego_support <- safe_rbind(lapply(
  split(ego_support, ego_support$anchor_hog),
  function(rows) {
    rows <- rows[order(
      -abs(rows$difference),
      -pmax(rows$support.Annual, rows$support.Perennial),
      rows$module_hog
    ), ]
    utils::head(rows, 12L)
  }
))

make_ego_plot <- function(hogs) {
  support <- ego_support[ego_support$anchor_hog %in% hogs, ]
  plot_rows <- list()
  edge_rows <- list()
  for (hog in hogs) {
    rows <- support[support$anchor_hog == hog, , drop = FALSE]
    rows <- rows[order(rows$module_hog), ]
    angle <- seq(0, 2 * pi, length.out = nrow(rows) + 1L)[
      seq_len(nrow(rows))
    ]
    rows$x <- cos(angle)
    rows$y <- sin(angle)
    label <- profile_meta$display_label[
      match(hog, profile_meta$anchor_hog)
    ]
    for (group in c("Annual", "Perennial")) {
      support_column <- paste0("support.", group)
      group_rows <- rows
      group_rows$life_cycle <- group
      group_rows$candidate <- label
      plot_rows[[length(plot_rows) + 1L]] <- group_rows
      present <- group_rows[[support_column]] > 0
      if (any(present)) {
        edge_rows[[length(edge_rows) + 1L]] <- data.frame(
          candidate = label,
          life_cycle = group,
          x = 0,
          y = 0,
          xend = group_rows$x[present],
          yend = group_rows$y[present],
          support = group_rows[[support_column]][present]
        )
      }
    }
  }
  nodes <- safe_rbind(plot_rows)
  edges <- safe_rbind(edge_rows)
  ggplot() +
    geom_segment(
      data = edges,
      aes(x = x, y = y, xend = xend, yend = yend, linewidth = support),
      colour = "#9aa5ab",
      alpha = 0.65,
      lineend = "round"
    ) +
    geom_point(
      data = nodes,
      aes(x = x, y = y, fill = bias),
      shape = 21,
      colour = "white",
      stroke = 0.4,
      size = 4
    ) +
    geom_point(
      data = unique(nodes[c("candidate", "life_cycle")]),
      aes(x = 0, y = 0),
      shape = 21,
      fill = "#243746",
      colour = "white",
      size = 6
    ) +
    geom_text(
      data = nodes,
      aes(x = 1.13 * x, y = 1.13 * y, label = sub("HOG:", "", module_hog)),
      size = 2.1,
      check_overlap = TRUE
    ) +
    facet_grid(life_cycle ~ candidate) +
    scale_fill_manual(values = c(
      "Annual-enriched" = annual_colour,
      "Perennial-enriched" = perennial_colour,
      "Shared or weak" = neutral_colour
    )) +
    scale_linewidth_continuous(range = c(0.4, 2.4), limits = c(1, 4)) +
    coord_equal(xlim = c(-1.35, 1.35), ylim = c(-1.35, 1.35)) +
    labs(fill = "Neighbour bias", linewidth = "Species support") +
    theme_void(base_size = 9) +
    theme(
      legend.position = "top",
      strip.text.x = element_text(size = 7.5)
    )
}

ego_pages <- split(ego_hogs, ceiling(seq_along(ego_hogs) / 4L))
for (page in ego_pages) {
  print(make_ego_plot(page))
}
```

### Sortable ego-network evidence

```{r}

ego_table <- ego_support
ego_table$candidate <- ego_table$anchor_hog
ego_table$neighbour <- ego_table$module_hog
ego_table <- ego_table[c(
  "candidate", "neighbour", "support.Annual",
  "support.Perennial", "difference", "bias"
)]

gt::gt(ego_table) |>
  gt::cols_label(
    candidate = "Candidate",
    neighbour = "Neighbour HOG",
    support.Annual = "Annual support",
    support.Perennial = "Perennial support",
    difference = "\u0394 support",
    bias = "Bias"
  ) |>
  gt::data_color(
    columns = "difference",
    palette = c(perennial_colour, "#f7f7f7", annual_colour),
    domain = c(-4, 4)
  ) |>
  gt::opt_interactive(
    use_sorting = TRUE,
    use_search = TRUE,
    use_filters = TRUE,
    use_compact_mode = TRUE,
    use_page_size_select = TRUE,
    page_size_default = 20
  )
```

## Annotation and external evidence

Database links are provided only when the corresponding identifier is
available. An OrthoDB search link is retained for every candidate HOG.

```{r}

annotation <- unique(candidates[c(
  "anchor_hog", "hotspot_direction", "is_hotspot",
  "is_recurrent_regulator", "annotation", "bdis_selected_gene"
)])
annotation <- merge(
  annotation,
  literature,
  by.x = "anchor_hog",
  by.y = "hog",
  all.x = TRUE,
  sort = FALSE
)
annotation$bdis_link <- ifelse(
  !is.na(annotation$bdis_gene) & nzchar(annotation$bdis_gene),
  paste0(
    "[", annotation$bdis_gene, "](",
    "https://plants.ensembl.org/Multi/Search/Results?q=",
    utils::URLencode(annotation$bdis_gene, reserved = TRUE),
    ")"
  ),
  ifelse(
    !is.na(annotation$bdis_selected_gene) &
      nzchar(annotation$bdis_selected_gene),
    paste0(
      "[", annotation$bdis_selected_gene, "](",
      "https://plants.ensembl.org/Multi/Search/Results?q=",
      utils::URLencode(
        annotation$bdis_selected_gene,
        reserved = TRUE
      ),
      ")"
    ),
    ""
  )
)
annotation$uniprot_link <- ifelse(
  !is.na(annotation$bdis_uniprot) & nzchar(annotation$bdis_uniprot),
  paste0(
    "[", annotation$bdis_uniprot, "](",
    "https://www.uniprot.org/uniprotkb/",
    annotation$bdis_uniprot,
    "/entry)"
  ),
  ""
)
annotation$orthodb_link <- ifelse(
  !is.na(annotation$orthodb_viridiplantae) &
    nzchar(annotation$orthodb_viridiplantae),
  paste0(
    "[", annotation$orthodb_viridiplantae, "](",
    "https://www.orthodb.org/?query=",
    annotation$orthodb_viridiplantae,
    ")"
  ),
  ""
)
annotation$doi_link <- ifelse(
  !is.na(annotation$doi) & nzchar(annotation$doi),
  paste0(
    "[DOI](",
    "https://doi.org/",
    sub(";.*$", "", annotation$doi),
    ")"
  ),
  ""
)
annotation$resolved <- first_nonempty(
  annotation$resolved_family,
  annotation$annotation
)
annotation_table <- annotation[c(
  "anchor_hog", "bdis_link", "uniprot_link", "orthodb_link",
  "hotspot_direction", "is_recurrent_regulator", "resolved",
  "orthology_status", "doi_link", "interpretation"
)]

gt::gt(annotation_table) |>
  gt::cols_label(
    anchor_hog = "HOG",
    bdis_link = "B. distachyon",
    uniprot_link = "UniProt",
    orthodb_link = "OrthoDB",
    hotspot_direction = "Signal",
    is_recurrent_regulator = "Regulator",
    resolved = "Resolved annotation",
    orthology_status = "Orthology",
    doi_link = "Literature",
    interpretation = "Interpretation"
  ) |>
  gt::fmt_markdown(columns = c(
    "bdis_link", "uniprot_link", "orthodb_link", "doi_link"
  )) |>
  gt::fmt_tf(columns = "is_recurrent_regulator") |>
  gt::opt_interactive(
    use_sorting = TRUE,
    use_search = TRUE,
    use_filters = TRUE,
    use_compact_mode = TRUE,
    use_page_size_select = TRUE,
    page_size_default = 15
  )
```

## Complete candidate matrix

This table preserves the principal phase, abundance, amplitude, shape, and
anchor–module lag measurements. Use the column filters to isolate a tissue,
evidence pattern, or direction.

```{r}

detail <- candidates[c(
  "anchor_hog", "tissue", "pattern_label", "phase_label",
  "anchor_phase_difference", "anchor_pair_agreement",
  "anchor_bootstrap_support",
  "anchor_adjusted_phase_median_annual_minus_perennial",
  "module_phase_difference", "module_pair_agreement",
  "anchor_mean_median_annual_minus_perennial",
  "module_mean_median_annual_minus_perennial",
  "anchor_amplitude_median_annual_minus_perennial",
  "module_amplitude_median_annual_minus_perennial",
  "anchor_shape_median_annual_minus_perennial",
  "module_shape_median_annual_minus_perennial",
  "phase_lag_median_annual_minus_perennial",
  "hotspot_direction", "is_hotspot", "is_recurrent_regulator",
  "annotation", "database_links"
)]

gt::gt(detail) |>
  gt::cols_label(
    anchor_hog = "HOG",
    tissue = "Tissue",
    pattern_label = "Pattern",
    phase_label = "Direction",
    anchor_phase_difference = "Anchor phase",
    anchor_pair_agreement = "Anchor pairs",
    anchor_bootstrap_support = "Bootstrap",
    anchor_adjusted_phase_median_annual_minus_perennial =
      "Adjusted anchor phase",
    module_phase_difference = "Module phase",
    module_pair_agreement = "Module pairs",
    anchor_mean_median_annual_minus_perennial = "Anchor mean",
    module_mean_median_annual_minus_perennial = "Module mean",
    anchor_amplitude_median_annual_minus_perennial = "Anchor amplitude",
    module_amplitude_median_annual_minus_perennial = "Module amplitude",
    anchor_shape_median_annual_minus_perennial = "Anchor shape shift",
    module_shape_median_annual_minus_perennial = "Module shape shift",
    phase_lag_median_annual_minus_perennial = "Anchor\u2212module lag",
    hotspot_direction = "Original signal",
    is_hotspot = "Hotspot",
    is_recurrent_regulator = "Regulator",
    annotation = "Annotation",
    database_links = "Databases"
  ) |>
  gt::fmt_markdown(columns = "database_links") |>
  gt::fmt_number(
    columns = where(is.numeric),
    decimals = 2
  ) |>
  gt::fmt_percent(columns = "anchor_bootstrap_support", decimals = 0) |>
  gt::fmt_tf(columns = c("is_hotspot", "is_recurrent_regulator")) |>
  gt::opt_interactive(
    use_sorting = TRUE,
    use_search = TRUE,
    use_filters = TRUE,
    use_resizers = TRUE,
    use_compact_mode = TRUE,
    use_page_size_select = TRUE,
    page_size_default = 15,
    height = "700px"
  )
```

## Reading the screen

- **Coordinated shift** means candidate and recurrent module passed the same
  recurrence, bootstrap, and empirical-effect criteria in the same direction.
- **Anchor only** and **module only** distinguish gene timing from program
  timing.
- **Adjusted phase** subtracts the species-, tissue-, and layer-specific median
  of all 336 eligible anchors. It controls broad transcriptome timing trends
  but is not a substitute for an independently measured developmental stage.
- **Ego-network support** is the number of species out of four in which a HOG
  is a direct neighbour of the candidate. It does not imply activation,
  repression, or physical binding.
- The five-point shape alignment is deliberately bounded to one sampling
  interval in either direction. Treat boundary estimates as coarse.
